### 2016-09-28 manuscript/SupplementaryMaterial.Rmd
Reformatting for GiM and adressing reviewers comments

---
title: "Supplementary Material - Using high-resolution variant frequencies to empower clinical genome interpretation"
output: 
  word_document:
    reference_docx: template2.docx

bibliography: "suppReferences.bib"
csl: "nature.csl"
---

```{r, echo=FALSE, include=FALSE}
#read all scripts in the R directory, to load functions
sourceDir <- function(path, trace = TRUE, ...) {
    for (nm in list.files(path, pattern = "\\.[RrSsQq]$")) {
       if(trace) cat(nm,":")           
       source(file.path(path, nm), ...)
       if(trace) cat("\n")
    }
}

#load functions
#sourceDir(file.path("..","R"))
source("../R/setup.R")
```

## Supplementary Table S1

Variants previously reported as causative of HCM either in ClinVar, or in a clinical series of 6179 HCM cases, but that were observed in ExAC above the maximum tolerated allele count for HCM (AC>9 globally) were manually curated according to the ACMG guidelines for interpretation and classification on sequence variants. *These variants were previously categorised as Likely Pathogenic in the LMM and Oxford HCM cohorts.

```{r}
read.delim("../data-raw/SupplementaryTable1.txt") %>% kable
```

## Suplementary Methods

### Curation of a high frequency PCD variant

_NME8_ NM\_016616.4:c.271-27C>T which is reported as Pathogenic in ClinVar is found in 2306/120984 ExAC individuals. This variant was initially reported as pathogenic on the basis of two compound heterozygous cases when specifically searching for NME8 variants in a set of patients, and was found in 2/196 control chromosomes `r citep("10.1073/pnas.0611405104")`. We further note that _NME8_ has not otherwise been associated with PCD and shows no evidence of missense nor loss-of-function constraint and that this splice variant affects a non-canonical transcript. During our curation exercise we found that this variant meets none of the ACMG criteria for assertions of pathogenicity, and therefore we reclassified it as a VUS.

### Dealing with penetrance

It is often difficult to obtain accurate penetrance information for reported variants, and it is also difficult to know what degree of penetrance to expect or assume in newly discovered pathogenic variants. In this work we uniformly apply a value of 50% penetrance (i.e. assuming variant penetrance is no lower than 50%) equivalent to that reported for our HCM example variant, and the lowest we found reported for any of our examples or reported accross our studied disorders. We recognise several other approaches that can be used to deal with the issues of penetrance, these include: setting a penetrance level equivalent to the minimum that is 'clinically actionable' for a disorder; lowering the penetrance if reduced penetrance is expected in a family; or using a two tiered approach, initially searching for a high-moderate penetrance variant but allowing for a lower-penetrance variant in a second pass. We believe that the ease of re-calculating our "maximum credible population allele frequency" lends itself to any of these approaches. We provide an online calculator to facilitate the exploration of these parameters (<https://jamesware.shinyapps.io/alleleFrequencyApp/>). If there are large case and control populations for a disease and the diease prevalence is known, we can use these to estimate penetrance `r citep("10.1126/scitranslmed.aad5169")`.

### Treatment of singletons and other populations

It is worth considering whether a single observation in a reference sample should ever be treated as incompatible with disease. Using the approach outlined above, it can be inferred that an ExAC AC=1 would be considered incompatible with a true population allele frequency <2.9x10^-6^ (with 95% confidence). For a penetrant disease with a prevalence of 1:1,000,000, the probability of observing a specific causative allele in ExAC is <0.01, even if the disease is genetically homogeneous with just one causative variant. In practice however, we feel that there are few, if any, diseases that are extremely rare yet have sufficiently well-characterized genetic architecture to discard singleton variants from a reference sample. Therefore, for singletons (variants observed exactly once in ExAC), we set the filtering allele frequency to zero.

We also note that occasionally a variant is seen in individuals falling under the Finnish or "Other" population categories in ExAC, and is a singleton or absent in all five continental populations. For these variants, the filtering allele frequency is set to zero. Because the Finnish are a bottlenecked population, disease-causing alleles may reach frequencies that would be impossible in large outbred populations. Similarly, because we have not assigned ancestry for the "Other" individuals, it is difficult to assess the population frequency of variants seen only in this set of individuals. Users are left to judge whether variants that  would not be filtered on the basis of frequency in the five continental populations, but that are recurrent in Finnish or "Other" populations, should be removed from consideration according to the specific circumstances.

### Description of the filtering allele frequency

We define the "filtering allele frequency" for a variant, or `af_filter`, as the highest true population allele frequency for which the upper bound of the 95% confidence interval of allele count under a Poisson distribution is still less than the variant's observed allele count in the reference sample. It functions as equivalent to a lower bound estimate for the true allele frequency of an observed variant: if the filtering allele frequency of a variant is at or above the maximum credible allele frequency for a disease, then the variant is considered too common to be causative of the disease.

Consider, for example, a variant with an observed AC=3 and AN=100,000. If a user's maximum credible allele frequency for their disease is 1 in 100,000, then this variant should be kept in consideration as potentially pathogenic, because the upper bound of the Poisson 95%CI is AC=3. On the other hand, if the user's credible tolerated allele frequency is 1 in 200,000 then this variant should be filtered out, as the 95%CI upper bound is only AC=2. We define `af_filter` as the highest AF value for which a variant should be filtered out.

In the example, the highest allele frequency that gives a 95%CI AC of 2 when AN=100,000 is approximately 8.17e-6. Instead of solving exactly for such values, which would require solving the inverse cumulative distribution function of the Poisson distribution, we derive a numerical approximation in two steps:

1. For each variant in consideration, we use R's `uniroot` function to find an AF value (though not necessarily the highest AF value) for which the 95%CI AC is one less than the observed AC.
2. We then loop, incrementing by units of millionths, and return the highest AF value that still gives a 95%CI AC less than the observed AC.

In order to pre-compute `af_filter` values for all of ExAC (verson 0.3.1), we apply this procedure to the AC and AN values for each of the five major continental populations in ExAC, and take the highest result from any population. Usually, this is from the population with the highest nominal allele frequency. However, because the tightness of a 95% confidence interval in the Poisson distribution depends upon sample size, the stringency of the filter depends upon the allele number (AN). The stringency of the filter therefore varies appropriately according the the size of the sub-population in which the variant is observed, and sequencing coverage at that site, and `af_filter` is occasionally derived from a population other than the one with the highest nominal allele frequency.

For this analysis, we used adjusted AC and AN, meaning variant calls with GQ≥20 and DP≥10.

##Supplemental references

```{r, echo=FALSE, message=FALSE}
write.bibtex(file="suppReferences.bib")
```

### 2016-09-28 manuscript/manuscript.Rmd
Reformatting for GiM and adressing reviewers comments

---
title: "Using high-resolution variant frequencies to empower clinical genome interpretation"
output:
#  html_document
  word_document:
    reference_docx: newTemplate.docx

bibliography: "references.bib"
csl: "nature.csl"
---

```{r, echo=FALSE, include=FALSE}
#load functions
source(file.path("..","R","setup.R"))
source(file.path("..","R","sumariseCaseSeriesVariants.R"))
#create figures
source(file.path("..","R","Figures.R"))
```

## Short running title: Using variant frequencies to empower genome interpretation

```{r authors, results='asis'}
read.delim("../data-raw/authors.txt") %>%
  formatAuthors
```

Corespondance should be addressed to:
Dr James Ware, National Heart and Lung Institute, Imperial College London, London, Tel: Fax: Email: j.ware@imperial.ac.uk

## ABSTRACT

**Purpose**: Whole exome and genome sequencing have transformed the discovery of genetic variants that cause human Mendelian disease, but discriminating pathogenic from benign variants remains a daunting challenge. Rarity is recognised as a necessary, although not sufficient, criterion for pathogenicity, but frequency cutoffs used in Mendelian analysis are often arbitrary and overly lenient. Recent very large reference datasets, such as the Exome Aggregation Consortium (ExAC), provide an unprecedented opportunity to obtain robust frequency estimates even for very rare variants.

**Methods**: We present a statistical framework for the frequency-based filtering of candidate disease-causing variants, accounting for disease prevalence, genetic and allelic heterogeneity, inheritance mode, penetrance, and sampling variance in reference datasets.

**Results**: Using the example of cardiomyopathy, we show that our approach reduces by two-thirds the number of candidate variants under consideration in the average exome, and identifies 43 variants previously reported as pathogenic that can now be reclassified.

**Conclusion**: We outline a statistically robust framework for assessing whether a variant is 'too common' to be causative for a Mendelian disorder of interest. We present precomputed allele frequency cutoffs for all variants in the ExAC dataset.

Key words:

## INTRODUCTION

Whole exome and whole genome sequencing have been instrumental in identifying causal variants in Mendelian disease patients `r citep("10.1016/j.ajhg.2015.06.009")`. As every individual harbors ~12,000-14,000 predicted protein-altering variants `r citep("10.1038/nature19057")`, distinguishing disease-causing variants from benign bystanders is perhaps the principal challenge in contemporary clinical genetics. A variant's low frequency in, or absence from, reference databases is now recognised as a necessary, but not sufficient, criterion for variant pathogenicity `r citep(c("10.1038/gim.2015.30","10.1038/nature13127"))`. The recent availability of very large reference databases, such as the Exome Aggregation Consortium (ExAC) `r citep("10.1038/nature19057")` dataset, which has characterised the population allele frequencies of 10 million genomic variants through the analysis of exome sequencing data from over 60,000 humans, provides an opportunity to obtain robust frequency estimates even for rare variants, improving the theoretical power for allele frequency filtering in Mendelian variant discovery efforts.

In practice, there exists considerable ambiguity around what allele frequency should be considered "too common", with the lenient values of 1% and 0.1% often invoked as conservative frequency cutoffs for recessive and dominant diseases respectively `r citep("10.1038/nrg3031")`. Population genetics, however, dictates that severe disease-causing variants must be much rarer than these cutoffs, except in cases of bottlenecked populations, balancing selection, or other special circumstances `r citep(c("10.1093/molbev/msp190","10.1073/pnas.1322563111"))`.

It is intuitive that when assessing a variant for a causative role in a dominant Mendelian disease, the frequency of a variant in a reference sample, not selected for the condition, should not exceed the prevalence of the condition `r citep(c("10.1016/j.ajhg.2016.03.024","10.1126/scitranslmed.aad5169"))`. This rule must, however, be refined to account for different inheritance modes, genetic and allelic heterogeneity, and reduced penetrance. In addition, for rare variants, estimation of true population allele frequency is clouded by considerable sampling variance, even in the largest samples currently available. These limitations have encouraged the adoption of very lenient approaches when filtering variants by allele frequency `r citep(c("10.1016/S0140-6736(14)61705-0","10.1038/ng.3304"))`, and recognition that more stringent approaches that account for disease-specific genetic architecture are urgently needed`r citep("10.1016/j.ajhg.2016.03.024")`.

Here we present a statistical framework for assessing whether rare variants are sufficiently rare to cause penetrant Mendelian disease, while accounting for both architecture and sampling variance in observed allele counts. We demonstrate that allele frequency cutoffs well below 0.1% are justified for a variety of human disease phenotypes and that such filters can remove an additional two-thirds of variants from consideration when compared to traditionally lenient frequency cutoffs. We present pre-computed allele frequency filtering values for all variants in the Exome Aggregation Consortium database, which are now available through the ExAC data browser and for download, to assist others in applying our framework.

## METHODS

### Calculating maximum tolerated allele counts

The maximum frequency of a dominant disease-causing variant in the population was defined as:  
*maximum credible population AF = prevalence* $\times$ *maximum allelic contribution* $\times$ *1/penetrance*

Disease prevalence estimates were obtained from the literature and taken as the highest value reported. A variant penetrance of 0.5 was used for all analyses, as penetrance estimates for individual variants are not widely available. This corresponds to the reported penetrance of the HCM variant used to illustrate our approach `r citep("10.1161/CIRCRESAHA.109.216291")` and is the minimum found when researching other variants/disorders.

Determination of the maximum allelic contribution (a measure of heterogeneity) is described in the text.

The maximum tolerated allele count (AC) was computed as the AC occurring at the upper bound of the one-tailed 95% confidence interval (95%CI AC) for the established maximum credible allele frequency, given the observed allele number (AN). Since the population is drawn without replacement, this would strictly be a hypergeometric distribution, but this can be modeled as binomial as the sample is much smaller than the population from which it is drawn. For ease of computation, we approximate this with a Poisson distribution. In R, this is implemented as `max_ac = qpois(quantile_limit,an*af)`, where `max_ac` is the 95%CI AC, `quantile_limit` is 0.95 (for a one-sided 95%CI), `an` is the observed allele number, and `af` is the maximum credible population allele frequency.

### Application to recessive diseases

The prevalence of a recessive condition can be related to the allele frequency of causative variants by:  
*Prevalence =* $\sum$ *(allele frequency of causative alleles in each contributing gene)^2^ x penetrance*  
approximating to:   
*Prevalence = (combined frequency of causative alleles in gene)^2^ x (number of similar genes) x penetrance*  
and expanding to:  
*Prevalence = (max individual allele frequency x 1/maximum allelic contribution)^2^ x 1/maximum genetic contribution x penetrance*

where *maximum genetic contribution* represents the proportion of all cases that are attributable to the gene under evaluation, and *maximum allelic contribution* represents the proportion of cases attributable to that gene that are attributable to an individual variant. The maximum frequency of a recessive disease causing variant in the population was therefore defined as:  
*max credible allele frequency* = $\sqrt(prevalence) \times$ *maximum allelic contribution* $\times \sqrt(maximum genetic contribution) \times 1/\sqrt(penetrance)$

### Pre-computing filtering allele frequency values for ExAC

In order to pre-compute `af_filter` values for all of ExAC (verson 0.3.1), we apply a two-step approach to the AC and AN values for each of the five major continental populations, and take the highest result from any population.

1. We use R's `uniroot` function to find an AF value (though not necessarily the highest AF value) for which the 95%CI AC is one less than the observed AC.
2. We loop, incrementing by units of millionths, and return the highest AF value that still gives a 95%CI AC less than the observed AC.

We used adjusted AC and AN, meaning variant calls with GQ≥20 and DP≥10. More explanation can be found in **Supplementary Methods**.

#### Simulated Mendelian variant discovery analysis

To simulate Mendelian variant discovery, we randomly selected 100 individuals from each of five major continental populations and filtered their exomes against filtering allele frequencies derived from the remaining 60,206 ExAC individuals. The subset of individuals was the same as that previously reported `r citep("10.1038/nature19057")`.
Predicted protein-altering variants are defined as missense and equivalent (including in-frame indels, start lost, stop lost, and mature miRNA-altering), and protein-truncating variants (nonsense, essential splice site, and frameshift).

### Variant curation

We utilized the July 9, 2015 release of ClinVar, extracting variants from XML and TXT releases into a single tab-delimited file through use of a Python implementation of vt normalize `r citep("10.1093/bioinformatics/btv112")`, as described previously `r citep("10.1038/nature19057")`. Only variants annotated as pathogenic and non-conflicted were investigated. ExAC counts were determined by matching on chromosome, position, reference, and alternate alleles. For all variants above the proposed maximum tolerated allele count for HCM, all HGMD annotated literature was reviewed and the level of evidence supporting disease pathogenicity was curated according to ACMG criteria `r citep("10.1038/gim.2015.30")`.

After curating variants above our calculated HCM threshold, the false discovery rate was 0/477 (0.000; 95%CI 0.000-0.008) and 1/559 (0.002; 95%CI 0.000-0.010) for the published HCM cohort and ClinVar data respectively.

### Calculating odds ratios for HCM variant burden

We used a cohort of 322 patients recruited to the Royal Brompton Hospital cardiac Biomedical Research Unit with diagnosis of HCM confirmed by cardiac MRI. These samples were sequenced using the IlluminaTruSight Cardio Sequencing Kit `r citep("10.1007/s12265-016-9673-5")` on theIlluminaMiSeq and NextSeq platforms. This study was subject to ethical approval (REC: 09/H0504/104+5) and informed consent was obtained for all subjects. The number of rare variants in _MYBPC3_, _MYH7_ and the six other sarcomeric genes associated with HCM (_TNNT2_, _TNNI3_, _MYL2_, _MYL3_, _TPM1_ and _ACTC1_) were calculated for this HCM cohort, and for reference population samples from ExAC. Case/control variant frequencies were calculated for all protein altering variants (frameshift, nonsense, splice donor/acceptor, missense and in-frame insertions/deletions), with frequencies and case/control odds ratios calculated separately for non-overlapping ExAC allele frequency bins with the following breakpoints: 1x10^-5^, 5x10^-5^, 1x10^-4^, 5x10^-4^ and 1x10^-3^. Odds Ratios were calculated as OR = (cases with variant / cases without variant) / (ExAC samples with variant / ExAC samples without variant) along with 95% confidence intervals. In the absence of sample-level genotype data for ExAC, the number of samples with a variant was approximated by the total number of variant alleles - i.e. assuming that each rare variant was found in a distinct sample.

## Data availability

All data required to reproduce these analyses is available at <https://github.com/ImperialCardioGenetics/frequencyFilter>. The manuscript was compiled in R, and source code for the analysis, figures and manuscript, are available at the same location. Curated variant interpretations are deposited in ClinVar *Accession & DOI to be added*. ExAC annotations are available at <exac.broadinstitute.org>. Our allele frequency calculator app is located at <https://jamesware.shinyapps.io/alleleFrequencyApp/>, and the source code available at <http://github.com/jamesware/alleleFrequencyApp>.

## RESULTS

#### *Defining the statistical framework*

For a penetrant dominant Mendelian allele to be disease causing, it cannot be present in the general population more frequently that the disease it causes. Furthermore, if the disease is genetically heterogeneous, it must not be more frequent than the proportion of cases attributable to that gene, or indeed to any single variant. We can therefore define the maximum credible population allele frequency (for a pathogenic allele) as:

*maximum credible population AF = prevalence x maximum allelic contribution x 1/penetrance*

where _maximum allelic contribution_ is the maximum proportion of cases potentially attributable to a single allele, a measure of heterogeneity.

We do not know the true population allele frequency of any variant, having only an observed allele frequency in a finite population sample. Moreover, confidence intervals around this observed frequency are problematic to estimate given our incomplete knowledge of the frequency spectrum of rare variants, which appears to be skewed towards very rare variants.  For instance, a variant observed only once in a sample of 10,000 chromosomes is much more likely to have a frequency < 1:10,000 than a frequency >1:10,000. `r citep("10.1038/nature19057")`

If we turn the problem around, and begin instead from allele frequency, specifying a maximum _true_ allele frequency value we are willing to consider in the population (using the equation above), then we can estimate the probability distribution for allele counts in a given sample size. This follows a binomial distribution, and can be satisfactorily approximated with a Poisson distribution (see **Methods**). This allows us to set an upper limit on the number of alleles in a sample that is consistent with a given population frequency.

Taking a range of cardiac disorders as exemplars, we use this framework to define the maximum credible allele frequency for disease-causing variants in each condition, and define and validate a set of maximum tolerated allele counts in the ExAC reference population sample.

#### *Application and validation in hypertrophic cardiomyopathy*

We illustrate our general approach using the dominant cardiac disorder hypertrophic cardiomyopathy (HCM), which has an estimated prevalence of 1 in 500 in the general population `r citep("10.1038/gim.2014.205")`. As there have been previous large-scale genetic studies of HCM, with series of up to 6,179 individuals `r citep(c("10.1038/gim.2014.205","10.1038/gim.2016.90"))`, we can make the assumption that no newly identified variant will be more frequent in cases that those identified to date (at least for well-studied ancestries). This allows us to define the maximum allelic contribution of any single variant to the disorder. In these large case series, the largest proportion of cases is attributable to the missense variant _MYBPC3_ c.1504C>T (p.Arg502Trp), found in 104/6179 HCM cases (1.7%; 95%CI 1.4-2.0%) `r citep(c("10.1038/gim.2014.205","10.1038/gim.2016.90"))`. We therefore take the upper bound of this proportion (0.02) as an estimate of the maximum allelic contribution in HCM (Table 1). Our maximum expected population allele frequency for this allele, assuming 50% penetrance as previously reported `r citep("10.1161/CIRCRESAHA.109.216291")`, is 1/500 x 1/2 (dividing prevalence per individual by the number of chromosomes per individual) x 0.02 x 1/0.5 = 4.0x10^-5^, which we take as the maximum credible population AF for any causative variant for HCM (Table 1).

To apply this threshold while remaining robust to chance variation in observed allele counts, we ask how many times a variant with population allele frequency of 4.0x10^-5^ can be observed in a random population sample of a given size. For a 5% error rate we take the 95th percentile of a poisson distribution with λ = expected allele count, which is given by: sample size (chromosomes) &times; expected population allele frequency (**Methods**).  For HCM this gives us a maximum tolerated allele count of 9, assuming 50% penetrance (or 5 for fully penetrant alleles), for variants genotyped in the full ExAC cohort (sample size=121,412 chromosomes). The _MYBPC3_:c.1504C>T variant is observed 3 times in ExAC (freq=2.49x10^-5^; Table 1).

To facilitate these calculations, we have produced an online calculator (<https://jamesware.shinyapps.io/alleleFrequencyApp/>) that will compute maximum credible population allele frequency and maximum sample allele count for a user-specified genetic architecture, and conversely allow users to dynamically explore what genetic architecture(s) might be most compatible with an observed variant having a causal role in disease.

```{r}
summariseCaseSeriesVariants()
```

To assess these thresholds empirically, we explored the ExAC allele frequency spectrum of `r myVariantsN` distinct autosomal variants identified in 6179 recently published HCM cases referred for diagnostic sequencing, and individually assessed and reported according to international guidelines`r citep(c("10.1038/gim.2014.205","10.1038/gim.2016.90"))`. 
`r myPassPath`/`r myAllPath` (`r signif(100*myPassPath/myAllPath,3)`%) variants reported as ‘Pathogenic’ or ‘Likely Pathogenic’ fell below our threshold (Figure 1), including all variants with a clear excess in cases. `r myAbsentPath` of these variants are absent from ExAC.  The `r myFailPath` variants historically classified as ‘Likely Pathogenic’, but prevalent in ExAC in this analysis, were reassessed using contemporary ACMG criteria: there was no strong evidence in support of pathogenicity, and they were reclassified in light of these findings (Table S1).
This analysis identifies `r myFailVus`/`r myAllVus` (`r signif(100*myFailVus/myAllVus,3)`%) VUS that are unlikely to be truly causative for HCM.

The above analysis applied a single global allele count limit of 9 for HCM, however, as allele frequencies differ between populations, filtering based on frequencies in individual populations may provide greater power `r citep("10.1038/nature19057")`. For example, a variant relatively common in any one population is unlikely pathogenic, even if rare in other populations, provided the disease prevalence and architecture is consistent across populations. We therefore compute a maximum tolerated AC for each distinct sub-population of our reference sample, and filter based on the highest allele frequency observed in any major continental population (see **Methods** and **Suplementary Note 3**).

To further validate this approach, we examined all 601 variants identified in ClinVar `r citep("10.1093/nar/gkt1113")` as "Pathogenic" or "Likely Pathogenic" and non-conflicted for HCM. 558 (93%) were sufficiently rare when assessed as described. 43 variants were insufficiently rare in at least one ExAC population, and were therefore re-curated. 42 of these had no segregation or functional data sufficient to demonstrate pathogenicity in the heterozygous state, and we would classify as VUS at most. The remaining variant (MYBPC3:c.3330+5G>C) had convincing evidence of pathogenicity, though with uncertain penetrance (see **Supplementary Methods**), and was observed twice in the African/African American ExAC population. This fell outside the 95% confidence interval for an underlying population frequency <4x10^-5^, but within the 99% confidence threshold: a single outlier due to stochastic variation is unsurprising given that these nominal probabilities are not corrected for multiple testing across 601 variants. In light of our updated assessment, 20 variants were reclassified as Benign/Likely Benign and 22 as VUS according to the American College for Medical Genetics and Genomics (ACMG) guidelines for variant interpretation `r citep("10.1038/gim.2015.30")` (Table S1).

#### *Extending this approach to other disorders*

This approach can be readily applied in diseases where large case series are available to assess the genetic and allelic architecture, such as the inherited cardiac conditions displayed in Table 1. In the absence of large case series, we must estimate the genetic architecture parameters by extrapolating from similar disorders and/or variant databases.

Where disease-specific variant databases exist, we can use these to help estimate the maximum allelic contribution in lieu of individual case series. For example, Marfan syndrome is a rare connective tissue disorder caused by variants in the _FBN1_ gene. The UMD-FBN1 database `r citep("10.1002/humu.10249")` contains 3077 variants in _FBN1_ from 280 references (last updated 28/08/14). The most common variant is in 30/3006 records (1.00%; 95CI 0.53-1.46%), which likely overestimates its contribution to disease if related individuals are not systematically excluded. Taking the upper bound of this frequency as our maximum allelic contribution, we derive a maximum tolerated allele count of 2 (Table 2). None of the five most common variants in the database are present in ExAC.

Where no mutation database exists, we can use what is known about similar disorders to estimate the maximum allelic contribution. For the cardiac conditions with large cases series in Table 1, the maximum proportion of cases attributable to any one variant is 6.7% (95CI 4.1-9.2%; PKP2:c.2146-1G>C found in 24/361 ARVC cases `r citep("10.1038/gim.2016.90")`). We therefore take the upper bound of this confidence interval (rounded up to 0.1) as an estimate of the maximum allelic contribution for other genetically heterogeneous cardiac conditions, unless we can find disease-specific evidence to alter it. For Noonan syndrome and Catecholaminergic Polymorphic Ventricular Tachycardia (CPVT - an inherited cardiac arrhythmia syndrome) with prevalences of 1 in 1000 `r citep("10.1016/S0140-6736(12)61023-X")` and 1 in 10,000 `r citep("10.1038/ejhg.2013.55")` respectively, this translates to maximum population frequencies of 5x10^-5^ and 5x10^-6^ and maximum tolerated ExAC allele counts of 10 and 2 (Table 2).

Finally, if the allelic heterogeneity of a disorder is not well characterised, it is conservative to assume minimal heterogeneity, so that the contribution of each gene is modelled as attributable to one allele, and the maximum allelic contribution is substituted by the maximum genetic contribution (i.e the maximum proportion of the disease attributable to single gene). For classic Ehlers-Danlos syndrome, up to 40% of the disease is caused by variation in the _COL5A1_ gene `r citep("10.1097/GIM.0b013e3181eed412")`. Taking 0.4 as our maximum allelic contribution, and a population prevalence of 1/20,000 `r citep("10.1097/GIM.0b013e3181eed412")` we derive a maximum tolerated ExAC AC of 5 (Table 2).

Here we have illustrated frequencies analysed at the level of the disease. In some cases this may be further refined by calculating distinct thresholds for individual genes, or even variants. For example, if there is one common founder mutation but no other variants that are recurrent across cases, then it would make sense to have the founder mutation as an exception to the calculated threshold.

#### *Application to recessive diseases*

So far we have considered diseases with a dominant inheritance model. Our framework is readily modified for application in recessive disease, and to illustrate this we consider the example of Primary Ciliary Dyskinesia (PCD), which has a prevalence of up to 1 in 10,000 individuals in the general population `r citep("10.1136/archdischild-2013-304831")`.

Intuitively, if one penetrant recessive variant were to be responsible for all PCD cases, it could have a maximum population frequency of $\sqrt(1/10000)$. The maximum frequency of a recessive disease-causing variant in the population can be more completely defined as:  
*max credible allele frequency* = $\sqrt(prevalence) \times$ *maximum allelic contribution* $\times \sqrt(maximum genetic contribution) \times 1/\sqrt(penetrance)$

where *maximum genetic contribution* represents the proportion of all cases that are attributable to the gene under evaluation, and *maximum allelic contribution* represents the proportion of cases attributable to that gene that are attributable to an individual variant (full derivation can be found in **Methods**).

We can refine our evaluation of PCD by estimating the maximum genetic and allelic contribution. Across previously published cohorts of PCD cases `r citep(c("10.1164/rccm.200603-370OC","10.1164/rccm.200601-084OC","10.1038/ng.2277"))`, *DNAI1* IVS1+2\_3insT was the most common variant with a total of 17/358 alleles (4.7% 95CI 2.5-7.0%). Given that ~9% of all patients with PCD have disease-causing variants in _DNAI1_ and the IVS1+2\_3insT variant is estimated to account for ~57% of variant alleles in _DNAI1_ `r citep("10.1164/rccm.200603-370OC")`, we can take these values as estimates of the maximum genetic and allelic contribution for PCD, yielding a maximum expected population AF of $\sqrt(1/10000) \times 0.57 \times \sqrt0.09 \times 1/\sqrt0.5 = 2.42 \times10$^-3^ This translates to a maximum tolerated ExAC AC of 322. *DNAI1* IVS1+2\_3insT is itself present at 56/121108 ExAC alleles (45/66636 non-Finnish European alleles). A single variant reported to cause PCD in ClinVar occurs in ExAC with AC > 332 (_NME8_ NM\_016616.4:c.271-27C>T; AC=2306/120984): our model therefore indicates that this variant frequency is too common to be disease-causing, and consistent with this we note that it meets none of the current ACMG criteria for assertions of pathogenicity, and have reclassified it as VUS (see **Supplementary Methods**).

#### *Pre-computing threshold values for the ExAC populations*

For each ExAC variant, we defined a "filtering allele frequency" that represents the threshold disease-specific "maximum credible allele frequency" at or below which the disease could not plausibly be caused by that variant. A variant with a filtering allele frequency ≥ the user-defined maximum credible allele frequency for the disease under consideration should be filtered, while a variant with a filtering allele frequency below the maximum credible remains a candidate (Figure 3). It is important to note that the filtering allele frequency is not desease speific and hence can be applied to any disease of interest. This value has been pre-computed for all variants in ExAC (see **Methods** and **Supplementary Methods**), and is available via the ExAC VCF and browser (http://exac.broadinstitute.org).

To assess the efficiency of our approach, we calculated the filtering allele frequency based on 60,206 exomes from ExAC and applied these filters to a simulated dominant Mendelian variant discovery analysis on the remaining 500 exomes (see **Methods**). Filtering at allele frequencies lower than 0.1% can substantially reduce the number of predicted protein-altering variants in consideration, with the mean number of variants per exome falling from 176 at a cutoff of 0.1% to 63 at a cutoff of 0.0001% (Figure 2a). Additionally, we compared the prevalence of variants in HCM genes in cases and controls across the allele frequency spectrum, and computed disease odds ratios for different frequency bins. The odds ratio for disease-association increases markedly at very low allele frequencies (Figure 2b) demonstrating that increasing the stringency of a frequency filter improves the information content of a genetic result.

## DISCUSSION

We have outlined a statistically robust framework for assessing whether a variant is 'too common' to be causative for a Mendelian disorder of interest. To our knowledge, there is currently no equivalent guidance on the use of variant frequency information, resulting in inconsistent thresholds across both clinical and research settings. Furthermore, though disease-specific thresholds are recommended`r citep("10.1016/j.ajhg.2016.03.024")`, in practice the same thresholds may be used across all diseases, even where they have widely differing genetic architectures and prevalences. We have shown the importance of applying stringent AF thresholds, in that many more variants can be removed from consideration, and the remaining variants have a much higher likelihood of being relevant. We also show, using HCM as an example, how lowering this threshold does not remove true dominant pathogenic variants.

In order to assist others in applying our framework, we have precomputed a 'filtering allele frequency' for all variants across the ExAC dataset. This is defined such that if the filtering allele frequency of a variant is at or above the user-defined "maximum credible population allele frequency" for the disease in question, then that variant is not a credible candidate (in other words, for any population allele frequency below the threshold value, the probability of the observed allele count in the ExAC sample is <0.05). Once a user has determined their "maximum credible population allele frequency", they may remove from consideration ExAC variants for which the filtering allele frequency is greater than or equal to than the chosen value.

Our method is designed to be complemenary to and used along side other gene and variant level methods to filter and prioritise candiate variants (e.g. gene level constraint`r citep("10.1038/ng.3050")`, amino acid conservation and missense prediction algorithms) along with segregation and functional data.

We recognize several limitations of our approach. First, the approach is limited by our understanding of the prevalence and genetic architecture of the disease in question: this characterisation will vary for different diseases and in different populations, though we illustrate approaches to estimation and extrapolation of parameters. In particular, we must be wary of extrapolating to or from less-well characterised populations that could harbour population-specific founder mutations. It is critical to define the genetic architecture in the population under study. We acknowledge that there is no substitute for knowning the genetic architecture of the condition of interest, and for some conditions this is not possible. It is the nature of any framework to rely on underlying assumptions, and we believe our method makes these explicit. If a parameter is particularly unclear, it is possible to define a range of architectures that fit with the observed variant. Secondly, it is often difficult to obtain accurate penetrance information for reported variants, and it is also difficult to know what degree of penetrance to expect or assume for newly discovered pathogenic variants (see **Supplementary Methods** for alternative approaches).

Thirdly, while we believe that ExAC is depleted of severe childhood inherited conditions, and not enriched for cardiomyopathies, it could be enriched relative to the general population for some Mendelian conditions, including Mendelian forms of common diseases such as diabetes or coronary disease that have been studied in contributing cohorts. Where this is possible, the maximum credible population allele frequency can be simply computed based on the estimated disease prevalence in the ExAC cohort, rather than the population prevalence. Finally, although the resulting allele frequency thresholds are more stringent than those previously used, they are likely to still be very lenient for many applications. For instance, we base our calculation on the most prevalent known pathogenic variant from a disease cohort. For HCM, for which more than 6,000 people have been sequenced, it is unlikely that any single newly identified variant, not previously catalogued in this large cohort, will explain a similarly large proportion of the disease as the most common causal variant, at least in well-studied populations. Future work may therefore involve modeling the frequency distribution of all known variants for a disorder, to further refine these thresholds.

The power of our approach is limited by currently available datasets. Increases in both the ancestral diversity and size of reference datasets will bring additional power to our method over time. We have avoided filtering on variants observed only once, because a single observation provides little information about true allele frequency (**Supplementary Methods**). A ten-fold increase in sample size, resulting from projects such as the US Precision Medicine Initiative, will separate vanishingly rare variants from those whose frequency really is ~1 in 100,000. Increased phenotypic information linked to reference datasets will also reduce limitations due to uncertain disease status, and improve prevalence estimates, adding further power to our approach.

## ACKNOWLEDGEMENTS

This work was supported by the Wellcome Trust (107469/Z/15/Z), the Medical Research Council (UK), the NIHR Biomedical Research Unit in Cardiovascular Disease at Royal Brompton & Harefield NHS Foundation Trust and Imperial College London, the Fondation Leducq (11 CVD-01), a Health Innovation Challenge Fund (HICF-R6-373) award from the Wellcome Trust and Department of Health, UK, and by the National Institute of Diabetes and Digestive and Kidney Diseases and the National Institute of General Medical Sciences of the NIH (awards U54DK105566 and R01GM104371). EVM is supported by the National Institutes of Health under a Ruth L. Kirschstein National Research Service Award (NRSA) NIH Individual Predoctoral Fellowship (F31) (award AI122592-01A1). AHO-L is supported by National Institutes of Health under Ruth L. Kirschstein National Research Service Award 4T32GM007748.

This publication includes independent research commissioned by the Health Innovation Challenge Fund (HICF), a parallel funding partnership between the Department of Health and Wellcome Trust. The views expressed in this work are those of the authors and not necessarily those of the Department of Health or Wellcome Trust.

## SUPPLEMENTARY INFORMATION

Supplementary table 1

Supplementary Methods:
- Curation of a high frequency PCD variant
- Dealing with penetrance
- Treatment of singletons and other populations
- Description of the filtering allele frequency

##REFERENCES

```{r, echo=FALSE, message=FALSE}
write.bibtex(file="references.bib")
```

##FIGURE AND TABLE LEGENDS

### Figure 1
> Plot of ExAC allele count (all populations) against case allele count for variants classified as VUS, Likely Pathogenic or Pathogenic in 6179 HCM cases. The dotted lines represent the maximum tolerated ExAC allele counts in HCM for 50% (dark blue) and 100% penetrance (light blue). Variants are colour coded according to reported pathogenicity. Where classifications from contributing laboratories were discordant the more conservative classification is plotted. The inset panel shows the full dataset, while the main panel expands the region of primary interest.

### Figure 2

> The clinical utility of stringent allele frequency thresholds. (a) The number of predicted protein-altering variants (definition in Methods) per exome as a function of the frequency filter applied. A one-tailed 95% confidence interval is used, meaning that variants were removed from consideration if their AC would fall within the top 5% of the Poisson probability distribution for the user's maximum credible AF (x axis). (b) The odds ratio for HCM disease-association against allele frequency. The prevalence of variants in HCM-associated genes (_MYH7_, _MYBPC3_ and other sarcomeric  (_TNNT2_, _TNNI3_, _MYL2_, _MYL3_, _TPM1_ and _ACTC1_, analysed collectively) in 322 HCM cases and 60,706 ExAC controls were compared for a range of allele frequency bins, and an odds ratio computed (see **Methods**). Data for each bin is plotted at the upper allele frequency cutoff. Error bars represent 95% confidence intervals. The probability that a variant is pathogenic is much greater at very low allele frequencies.

### Figure 3

> A flow diagram of our approach, applied to a dominant condition. First, a disease-level maximum credible population allele frequency is calculated, based on disease prevalence, heterogeneity and penetrance. This is then compared against the pre-calculated variant filtering allele frequency to determine if a variant is compatible with disease.

### Table 1  
> Details of the most prevalent pathogenic variants in case cohorts for five cardiac conditions. Shown along with the frequency in cases is the estimated population allele frequency (calculated as: case frequency x disease prevalence x 1/2 x 1/variant penetrance) and the observed frequency in the ExAC dataset. ^\*^As penetrance estimates for individual variants are not widely available, we have applied an estimate of 0.5 across all disorders (see **Supplementary Information**). HCM - hypertrophic cardiomyopathy; DCM - dilated cardiomyopathy; ARVC - arrhythmogenic right ventricular cardiomyopathy; LQTS - long QT syndrome. Case cohorts and prevalence estimates (taken as the highest value reported) were obtained from: HCM`r citep(c("10.1038/gim.2014.205","10.1038/gim.2016.90"))`, DCM`r citep(c("10.1038/gim.2014.205","10.1038/gim.2016.90","10.1038/nrcardio.2013.105"))`, ARVC`r citep(c("10.1038/gim.2016.90","10.1016/j.ijcard.2005.12.015"))`, LQTS`r citep(c("10.1016/j.hrthm.2009.05.021","10.1016/j.cjca.2012.07.847"))` and Brugada`r citep(c("10.1016/j.hrthm.2009.09.069","10.1016/j.hlc.2015.07.020"))`.

### Table 2
> Maximum credible population frequencies and maximum tolerated ExAC allele counts for variants causative of exemplar inherited cardiac conditions, assuming a penetrance of 0.5 throughout. CPVT - catecholaminergic polymorphic ventricular tachycardia; FH - familial hypercholesterolaemia. Prevalence estimates (taken as the highest value reported) were obtained from: Marfan`r citep("10.1016/S0140-6736(05)67789-6")`, Noonan`r citep("10.1016/S0140-6736(12)61023-X")`, CPVT`r citep("10.1038/ejhg.2013.55")` and classical Ehlers-Danlos`r citep("10.1097/GIM.0b013e3181eed412")`.

##TABLES

###Table 1
```{r}
read.delim("../data-raw/Table1.txt",check.names=F) %>% pander
```

###Table 2
```{r}
read.delim("../data-raw/Table2.txt",check.names=F) %>% pander
```

### 2016-09-29 manuscript/manuscript.Rmd
Converting to GiM style

@@
 bibliography: "references.bib"
-csl: "nature.csl"
+csl: "american-medical-association.csl"
 ---
@@
 
-## Short running title: Using variant frequencies to empower genome interpretation
+**Short running title: Allele frequency and genome interpretation**
 
@@
 Corespondance should be addressed to:
-Dr James Ware, National Heart and Lung Institute, Imperial College London, London, Tel: Fax: Email: j.ware@imperial.ac.uk
+Dr James Ware, Cardiovascular Genetics and Genomics, NHLI, RBH Campus, Imperial College London, London, SW3 6NP; Tel: +44 20 7352 8121; Fax: +44 20 7351 8146; Email: j.ware@imperial.ac.uk
+
+## CONFLICT OF INTEREST NOTIFICATION
+
+The authors have no conflicts of interest to declare.
 
@@
 
-Key words:
+Key words: Variant interpretation, Clinical Genomics, Allele frequency, Cardiomyopathy, Inherited Cardiovascular conditions, ExAC
 
@@
 
-## Data availability
+### Data availability
 

### 2016-10-03 manuscript/SupplementaryMaterial.Rmd
Cutting words

@@
 ## Suplementary Methods
+
+### The statistical model
+
+The maximum tolerated allele count (AC) was computed as the AC occurring at the upper bound of the one-tailed 95% confidence interval (95%CI AC) for the established maximum credible allele frequency, given the observed allele number (AN).
+Since the population is drawn without replacement, this would strictly be a hypergeometric distribution, but this can be modeled as binomial as the sample is much smaller than the population from which it is drawn.
+For ease of computation, we approximate this with a Poisson distribution.
+In R, this is implemented as `max_ac = qpois(quantile_limit,an*af)`, where `max_ac` is the 95%CI AC, `quantile_limit` is 0.95 (for a one-sided 95%CI), `an` is the observed allele number, and `af` is the maximum credible population allele frequency.
+
+### Defining maximum credible AF for recessive diseases
+
+The prevalence of a recessive condition can be related to the AF of causative variants by:  
+*Prevalence =* $\sum$ *(AF of causative alleles in each contributing gene)^2^ x penetrance*  
+approximating to:   
+*Prevalence = (combined frequency of causative alleles in gene)^2^ x (number of similar genes) x penetrance*  
+and expanding to:  
+*Prevalence = (max individual AF x 1/maximum allelic contribution)^2^ x 1/maximum genetic contribution x penetrance*
+
+where *maximum genetic contribution* represents the proportion of all cases that are attributable to the gene under evaluation, and *maximum allelic contribution* represents the proportion of cases attributable to that gene that are attributable to an individual variant.
+The maximum frequency of a recessive disease causing variant in the population was therefore defined as:
+*max credible AF* = $\sqrt(prevalence) \times$ *maximum allelic contribution* $\times \sqrt(maximum genetic contribution) \times 1/\sqrt(penetrance)$
 

### 2016-10-03 manuscript/manuscript.Rmd
Cutting words

@@
 As every individual harbors ~12,000-14,000 predicted protein-altering variants `r citep("10.1038/nature19057")`, distinguishing disease-causing variants from benign bystanders is perhaps the principal challenge in contemporary clinical genetics.
-A variant's low frequency in, or absence from, reference databases is now recognised as a necessary, but not sufficient, criterion for variant pathogenicity `r citep(c("10.1038/gim.2015.30","10.1038/nature13127"))`.
+A variant's low frequency in, or absence from, reference databases is recognised as a necessary, but not sufficient, criterion for variant pathogenicity `r citep(c("10.1038/gim.2015.30","10.1038/nature13127"))`.
 The recent availability of very large reference databases, such as the Exome Aggregation Consortium (ExAC) `r citep("10.1038/nature19057")` dataset, which has characterised the population allele frequencies of 10 million genomic variants through the analysis of exome sequencing data from over 60,000 humans, provides an opportunity to obtain robust frequency estimates even for rare variants, improving the theoretical power for allele frequency filtering in Mendelian variant discovery efforts.
@@
 
-The maximum frequency of a dominant disease-causing variant in the population was defined as:  
+The maximum alelle frequency (AF) of a dominant disease-causing variant in the population was defined as:  
 *maximum credible population AF = prevalence* $\times$ *maximum allelic contribution* $\times$ *1/penetrance*
-
-Disease prevalence estimates were obtained from the literature and taken as the highest value reported.
-A variant penetrance of 0.5 was used for all analyses, as penetrance estimates for individual variants are not widely available.
-This corresponds to the reported penetrance of the HCM variant used to illustrate our approach `r citep("10.1161/CIRCRESAHA.109.216291")` and is the minimum found when researching other variants/disorders.
 
@@
 
-The maximum tolerated allele count (AC) was computed as the AC occurring at the upper bound of the one-tailed 95% confidence interval (95%CI AC) for the established maximum credible allele frequency, given the observed allele number (AN).
-Since the population is drawn without replacement, this would strictly be a hypergeometric distribution, but this can be modeled as binomial as the sample is much smaller than the population from which it is drawn.
-For ease of computation, we approximate this with a Poisson distribution.
-In R, this is implemented as `max_ac = qpois(quantile_limit,an*af)`, where `max_ac` is the 95%CI AC, `quantile_limit` is 0.95 (for a one-sided 95%CI), `an` is the observed allele number, and `af` is the maximum credible population allele frequency.
-
-### Application to recessive diseases
-
-The prevalence of a recessive condition can be related to the allele frequency of causative variants by:  
-*Prevalence =* $\sum$ *(allele frequency of causative alleles in each contributing gene)^2^ x penetrance*  
-approximating to:   
-*Prevalence = (combined frequency of causative alleles in gene)^2^ x (number of similar genes) x penetrance*  
-and expanding to:  
-*Prevalence = (max individual allele frequency x 1/maximum allelic contribution)^2^ x 1/maximum genetic contribution x penetrance*
+For recessive conditions, the maximum AF is defined as:
+*max credible allele frequency* = $\sqrt(prevalence) \times$ *maximum allelic contribution* $\times \sqrt(maximum genetic contribution) \times 1/\sqrt(penetrance)$
 
 where *maximum genetic contribution* represents the proportion of all cases that are attributable to the gene under evaluation, and *maximum allelic contribution* represents the proportion of cases attributable to that gene that are attributable to an individual variant.
-The maximum frequency of a recessive disease causing variant in the population was therefore defined as:
-*max credible allele frequency* = $\sqrt(prevalence) \times$ *maximum allelic contribution* $\times \sqrt(maximum genetic contribution) \times 1/\sqrt(penetrance)$
+Full derivation can be found in **Supplementary Methods**.
+
+Disease prevalence estimates were obtained from the literature and taken as the highest value reported.
+Variant penetrance of 0.5 was used for all analyses, as penetrance estimates for individual variants are not widely available.
+This corresponds to the reported penetrance of the HCM variant used to illustrate our approach `r citep("10.1161/CIRCRESAHA.109.216291")` and is the minimum found when researching other variants/disorders.
+
+The maximum tolerated allele count (AC) was computed as the AC at the upper bound of the one-tailed 95% confidence interval (95%CI AC) of a Poisson distribution, for the established maximum credible AF, given the observed allele number (AN).
+In R: `max_ac = qpois(quantile_limit,an*af)`, where `max_ac` is the 95%CI AC, `quantile_limit` is 0.95 (for a one-sided 95%CI), `an` is the observed allele number, and `af` is the maximum credible population AF.
 
@@
 
-We used adjusted AC and AN, meaning variant calls with GQ≥20 and DP≥10. More explanation can be found in **Supplementary Methods**.
+More explanation can be found in **Supplementary Methods**.
+
+We used adjusted AC and AN, meaning variant calls with GQ≥20 and DP≥10.
 
@@
 
-We utilized the July 9, 2015 release of ClinVar, extracting variants from XML and TXT releases into a single tab-delimited file through use of a Python implementation of vt normalize `r citep("10.1093/bioinformatics/btv112")`, as described previously `r citep("10.1038/nature19057")`.
-Only variants annotated as pathogenic and non-conflicted were investigated.
+Pathogenic and non-conflicted variants from XML and TXT releases of ClinVar (July 9, 2015 release) were extracted using a Python implementation of vt normalize `r citep("10.1093/bioinformatics/btv112")`, as described previously `r citep("10.1038/nature19057")`.
 ExAC counts were determined by matching on chromosome, position, reference, and alternate alleles.
-For all variants above the proposed maximum tolerated allele count for HCM, all HGMD annotated literature was reviewed and the level of evidence supporting disease pathogenicity was curated according to ACMG criteria `r citep("10.1038/gim.2015.30")`.
+For all variants above the proposed maximum tolerated allele count for HCM, HGMD literature was reviewed and the level of evidence supporting pathogenicity was curated according to ACMG criteria `r citep("10.1038/gim.2015.30")`.
 
@@
 
-We used a cohort of 322 patients recruited to the Royal Brompton Hospital cardiac Biomedical Research Unit with diagnosis of HCM confirmed by cardiac MRI.
-These samples were sequenced using the IlluminaTruSight Cardio Sequencing Kit `r citep("10.1007/s12265-016-9673-5")` on theIlluminaMiSeq and NextSeq platforms.
-This study was subject to ethical approval (REC: 09/H0504/104+5) and informed consent was obtained for all subjects.
+322 HCM patients (confirmed by cardiac MRI) recruited to the Royal Brompton Hospital cardiac BRU were sequenced using the IlluminaTruSight Cardio Sequencing Kit `r citep("10.1007/s12265-016-9673-5")` on Illumina MiSeq and NextSeq platforms.
+This study had ethical approval (REC: 09/H0504/104+5) and informed consent was obtained for all subjects.
 The number of rare variants in _MYBPC3_, _MYH7_ and the six other sarcomeric genes associated with HCM (_TNNT2_, _TNNI3_, _MYL2_, _MYL3_, _TPM1_ and _ACTC1_) were calculated for this HCM cohort, and for reference population samples from ExAC.
-Case/control variant frequencies were calculated for all protein altering variants (frameshift, nonsense, splice donor/acceptor, missense and in-frame insertions/deletions), with frequencies and case/control odds ratios calculated separately for non-overlapping ExAC allele frequency bins with the following breakpoints: 1x10^-5^, 5x10^-5^, 1x10^-4^, 5x10^-4^ and 1x10^-3^.
-Odds Ratios were calculated as OR = (cases with variant / cases without variant) / (ExAC samples with variant / ExAC samples without variant) along with 95% confidence intervals.
-In the absence of sample-level genotype data for ExAC, the number of samples with a variant was approximated by the total number of variant alleles - i.e.
-assuming that each rare variant was found in a distinct sample.
+Variant frequencies were calculated for all protein altering variants (frameshift, nonsense, splice donor/acceptor, missense and in-frame insertions/deletions), with case/control odds ratios calculated separately for non-overlapping ExAC AF bins with the following breakpoints: 1x10^-5^, 5x10^-5^, 1x10^-4^, 5x10^-4^ and 1x10^-3^.
+Odds Ratios were calculated as OR = (cases with variant/cases without variant) / (ExAC samples with variant/ExAC samples without variant).
+In the absence of sample-level genotype data for ExAC, the number of samples with a variant was approximated by the total number of variant alleles (assuming each rare variant was from a distinct sample).
 
@@
 ExAC annotations are available at <exac.broadinstitute.org>.
-Our allele frequency calculator app is located at <https://jamesware.shinyapps.io/alleleFrequencyApp/>, and the source code available at <http://github.com/jamesware/alleleFrequencyApp>.
+Our allele frequency calculator app is located at <https://jamesware.shinyapps.io/alleleFrequencyApp/>, with source code available at <http://github.com/jamesware/alleleFrequencyApp>.
 
@@
 For each ExAC variant, we defined a "filtering allele frequency" that represents the threshold disease-specific "maximum credible allele frequency" at or below which the disease could not plausibly be caused by that variant.
-A variant with a filtering allele frequency ≥ the user-defined maximum credible allele frequency for the disease under consideration should be filtered, while a variant with a filtering allele frequency below the maximum credible remains a candidate (Figure 3).
-It is important to note that the filtering allele frequency is not desease speific and hence can be applied to any disease of interest.
+A variant with a filtering allele frequency≥the user-defined maximum credible allele frequency for the disease under consideration should be filtered, while a variant with a filtering allele frequency below the maximum credible remains a candidate (Figure 3).
+The filtering allele frequency is not desease speific and hence can be applied to any disease of interest.
 This value has been pre-computed for all variants in ExAC (see **Methods** and **Supplementary Methods**), and is available via the ExAC VCF and browser (http://exac.broadinstitute.org).
 
-To assess the efficiency of our approach, we calculated the filtering allele frequency based on 60,206 exomes from ExAC and applied these filters to a simulated dominant Mendelian variant discovery analysis on the remaining 500 exomes (see **Methods**).
-Filtering at allele frequencies lower than 0.1% can substantially reduce the number of predicted protein-altering variants in consideration, with the mean number of variants per exome falling from 176 at a cutoff of 0.1% to 63 at a cutoff of 0.0001% (Figure 2a).
+To assess the efficiency of our approach, we calculated the filtering allele frequency on 60,206 exomes from ExAC and applied these filters to a simulated dominant Mendelian variant discovery analysis on the remaining 500 exomes (see **Methods**).
+Filtering at allele frequencies lower than 0.1% substantially reduces the number of predicted protein-altering variants in consideration, with the mean number of variants per exome falling from 176 to 63 at cutoffs of 0.1% and 0.0001% respectively (Figure 2a).
 Additionally, we compared the prevalence of variants in HCM genes in cases and controls across the allele frequency spectrum, and computed disease odds ratios for different frequency bins.

### 2016-10-03 manuscript/manuscript.Rmd
Removed more words

@@
 
-In order to pre-compute `af_filter` values for all of ExAC (verson 0.3.1), we apply a two-step approach to the AC and AN values for each of the five major continental populations, and take the highest result from any population.
+In order to pre-compute `af_filter` values for all of ExAC (verson 0.3.1), we apply a two-step approach to the AC and AN values for each of the five major continental populations, and take the highest result from any population (more explanation in **Supplementary Methods**).
 
@@
 2. We loop, incrementing by units of millionths, and return the highest AF value that still gives a 95%CI AC less than the observed AC.
-
-More explanation can be found in **Supplementary Methods**.
 
@@
 For all variants above the proposed maximum tolerated allele count for HCM, HGMD literature was reviewed and the level of evidence supporting pathogenicity was curated according to ACMG criteria `r citep("10.1038/gim.2015.30")`.
-
-After curating variants above our calculated HCM threshold, the false discovery rate was 0/477 (0.000; 95%CI 0.000-0.008) and 1/559 (0.002; 95%CI 0.000-0.010) for the published HCM cohort and ClinVar data respectively.
 
@@
 
-In order to assist others in applying our framework, we have precomputed a 'filtering allele frequency' for all variants across the ExAC dataset.
+To assist others in applying our framework, we have precomputed a 'filtering allele frequency' for all variants across the ExAC dataset.
 This is defined such that if the filtering allele frequency of a variant is at or above the user-defined "maximum credible population allele frequency" for the disease in question, then that variant is not a credible candidate (in other words, for any population allele frequency below the threshold value, the probability of the observed allele count in the ExAC sample is <0.05).
@@
 
-We recognize several limitations of our approach.
-First, the approach is limited by our understanding of the prevalence and genetic architecture of the disease in question: this characterisation will vary for different diseases and in different populations, though we illustrate approaches to estimation and extrapolation of parameters.
+We recognise several limitations of our approach.
+First, the approach is limited by our understanding of the prevalence and genetic architecture of the disease in question: this characterisation will vary for different diseases and in different populations, though we illustrate approaches for estimation and extrapolation of parameters.
 In particular, we must be wary of extrapolating to or from less-well characterised populations that could harbour population-specific founder mutations.
@@
 
-Thirdly, while we believe that ExAC is depleted of severe childhood inherited conditions, and not enriched for cardiomyopathies, it could be enriched relative to the general population for some Mendelian conditions, including Mendelian forms of common diseases such as diabetes or coronary disease that have been studied in contributing cohorts.
+Thirdly, while we believe that ExAC is depleted of severe childhood inherited conditions, and not enriched for cardiomyopathies, it could be enriched relative to the general population for some conditions, including Mendelian forms of common diseases such as diabetes or coronary disease that have been studied in contributing cohorts.
 Where this is possible, the maximum credible population allele frequency can be simply computed based on the estimated disease prevalence in the ExAC cohort, rather than the population prevalence.
@@
 Increases in both the ancestral diversity and size of reference datasets will bring additional power to our method over time.
-We have avoided filtering on variants observed only once, because a single observation provides little information about true allele frequency (**Supplementary Methods**).
+We have avoided filtering on variants observed only once, because a single observation provides little information about true allele frequency (see **Supplementary Methods**).
 A ten-fold increase in sample size, resulting from projects such as the US Precision Medicine Initiative, will separate vanishingly rare variants from those whose frequency really is ~1 in 100,000.

### 2016-10-07 manuscript/manuscript.Rmd
More reformatting

@@
 Corespondance should be addressed to:
-Dr James Ware, Cardiovascular Genetics and Genomics, NHLI, RBH Campus, Imperial College London, London, SW3 6NP; Tel: +44 20 7352 8121; Fax: +44 20 7351 8146; Email: j.ware@imperial.ac.uk
+Dr James Ware, Cardiovascular Genetics and Genomics, National Heart and Lung Institute, Royal Brompton Hospital Campus, Imperial College London, London, SW3 6NP; Tel: +44 20 7352 8121; Fax: +44 20 7351 8146; Email: j.ware@imperial.ac.uk
 
@@
 
-**Results**: Using the example of cardiomyopathy, we show that our approach reduces by two-thirds the number of candidate variants under consideration in the average exome, and identifies 43 variants previously reported as pathogenic that can now be reclassified.
+**Results**: Using the example of cardiomyopathy, we show that our approach reduces by two-thirds the number of candidate variants under consideration in the average exome, without removing true pathogenic variants (false positive rate<0.001).
 
@@
 Here we present a statistical framework for assessing whether rare variants are sufficiently rare to cause penetrant Mendelian disease, while accounting for both architecture and sampling variance in observed allele counts.
-We demonstrate that allele frequency cutoffs well below 0.1% are justified for a variety of human disease phenotypes and that such filters can remove an additional two-thirds of variants from consideration when compared to traditionally lenient frequency cutoffs.
-We present pre-computed allele frequency filtering values for all variants in the Exome Aggregation Consortium database, which are now available through the ExAC data browser and for download, to assist others in applying our framework.
+We demonstrate that allele frequency cutoffs well below 0.1% are justified for a variety of human disease phenotypes and that such filters can remove an additional two-thirds of variants from consideration when compared to traditionally lenient frequency cutoffs, without discarding true pathogenic variants.
+We present pre-computed allele frequency filtering values for all variants in the Exome Aggregation Consortium database, for comparison with user-defined thresholds, which are now available through the ExAC data browser and for download, to assist others in applying our framework.
 
@@
 
-### Calculating maximum tolerated allele counts
+#### *Defining the statistical framework*
 
-The maximum alelle frequency (AF) of a dominant disease-causing variant in the population was defined as:  
+For a penetrant dominant Mendelian allele to be disease causing, it cannot be present in the general population more frequently that the disease it causes.
+Furthermore, if the disease is genetically heterogeneous, it must not be more frequent than the proportion of cases attributable to that gene, or indeed to any single variant.
+We can therefore define the maximum credible population allele frequency (for a pathogenic allele) as:
+
 *maximum credible population AF = prevalence* $\times$ *maximum allelic contribution* $\times$ *1/penetrance*
 
-Determination of the maximum allelic contribution (a measure of heterogeneity) is described in the text.
+where _maximum allelic contribution_ is the maximum proportion of cases potentially attributable to a single allele, a measure of heterogeneity.
 
@@
 This corresponds to the reported penetrance of the HCM variant used to illustrate our approach `r citep("10.1161/CIRCRESAHA.109.216291")` and is the minimum found when researching other variants/disorders.
+When the population frequency of a variant is known, we can use this to estimate prevalence `r citep("10.1126/scitranslmed.aad5169")`.
 
-The maximum tolerated allele count (AC) was computed as the AC at the upper bound of the one-tailed 95% confidence interval (95%CI AC) of a Poisson distribution, for the established maximum credible AF, given the observed allele number (AN).
+We do not know the true population allele frequency of any variant, having only an observed allele frequency in a finite population sample.
+Moreover, confidence intervals around this observed frequency are problematic to estimate given our incomplete knowledge of the frequency spectrum of rare variants, which appears to be skewed towards very rare variants.
+For instance, a variant observed only once in a sample of 10,000 chromosomes is much more likely to have a frequency < 1:10,000 than a frequency >1:10,000.
+`r citep("10.1038/nature19057")`
+
+If we turn the problem around, and begin instead from allele frequency, specifying a maximum _true_ allele frequency value we are willing to consider in the population (using the equation above), then we can estimate the probability distribution for allele counts in a given sample size.
+This follows a binomial distribution, and can be satisfactorily approximated with a Poisson distribution (see **Supplementary Methods**).
+This allows us to set an upper limit on the number of alleles in a sample that is consistent with a given population frequency.
+
+The maximum tolerated allele count (AC) was therefore computed as the AC at the upper bound of the one-tailed 95% confidence interval (95%CI AC) of a Poisson distribution, for the established maximum credible AF, given the observed allele number (AN).
 In R: `max_ac = qpois(quantile_limit,an*af)`, where `max_ac` is the 95%CI AC, `quantile_limit` is 0.95 (for a one-sided 95%CI), `an` is the observed allele number, and `af` is the maximum credible population AF.
@@
 
-322 HCM patients (confirmed by cardiac MRI) recruited to the Royal Brompton Hospital cardiac BRU were sequenced using the IlluminaTruSight Cardio Sequencing Kit `r citep("10.1007/s12265-016-9673-5")` on Illumina MiSeq and NextSeq platforms.
+322 HCM patients (confirmed by cardiac MRI) recruited to the NIHR Royal Brompton cardiovascular BRU were sequenced using the IlluminaTruSight Cardio Sequencing Kit `r citep("10.1007/s12265-016-9673-5")` on Illumina MiSeq and NextSeq platforms.
 This study had ethical approval (REC: 09/H0504/104+5) and informed consent was obtained for all subjects.
@@
 
-### Data availability
-
-All data required to reproduce these analyses is available at <https://github.com/ImperialCardioGenetics/frequencyFilter>.
-The manuscript was compiled in R, and source code for the analysis, figures and manuscript, are available at the same location.
-Curated variant interpretations are deposited in ClinVar *Accession & DOI to be added*.
-ExAC annotations are available at <exac.broadinstitute.org>.
-Our allele frequency calculator app is located at <https://jamesware.shinyapps.io/alleleFrequencyApp/>, with source code available at <http://github.com/jamesware/alleleFrequencyApp>.
-
 ## RESULTS
-
-#### *Defining the statistical framework*
-
-For a penetrant dominant Mendelian allele to be disease causing, it cannot be present in the general population more frequently that the disease it causes.
-Furthermore, if the disease is genetically heterogeneous, it must not be more frequent than the proportion of cases attributable to that gene, or indeed to any single variant.
-We can therefore define the maximum credible population allele frequency (for a pathogenic allele) as:
-
-*maximum credible population AF = prevalence x maximum allelic contribution x 1/penetrance*
-
-where _maximum allelic contribution_ is the maximum proportion of cases potentially attributable to a single allele, a measure of heterogeneity.
-
-We do not know the true population allele frequency of any variant, having only an observed allele frequency in a finite population sample.
-Moreover, confidence intervals around this observed frequency are problematic to estimate given our incomplete knowledge of the frequency spectrum of rare variants, which appears to be skewed towards very rare variants.
-For instance, a variant observed only once in a sample of 10,000 chromosomes is much more likely to have a frequency < 1:10,000 than a frequency >1:10,000.
-`r citep("10.1038/nature19057")`
-
-If we turn the problem around, and begin instead from allele frequency, specifying a maximum _true_ allele frequency value we are willing to consider in the population (using the equation above), then we can estimate the probability distribution for allele counts in a given sample size.
-This follows a binomial distribution, and can be satisfactorily approximated with a Poisson distribution (see **Methods**).
-This allows us to set an upper limit on the number of alleles in a sample that is consistent with a given population frequency.
-
-Taking a range of cardiac disorders as exemplars, we use this framework to define the maximum credible allele frequency for disease-causing variants in each condition, and define and validate a set of maximum tolerated allele counts in the ExAC reference population sample.
 
@@
 Our maximum expected population allele frequency for this allele, assuming 50% penetrance as previously reported `r citep("10.1161/CIRCRESAHA.109.216291")`, is 1/500 x 1/2 (dividing prevalence per individual by the number of chromosomes per individual) x 0.02 x 1/0.5 = 4.0x10^-5^, which we take as the maximum credible population AF for any causative variant for HCM (Table 1).
-
 To apply this threshold while remaining robust to chance variation in observed allele counts, we ask how many times a variant with population allele frequency of 4.0x10^-5^ can be observed in a random population sample of a given size.
@@
 
+We computed the maximum tolerated allele count for four additional cardiac disorders with available cases series.
+In all cases the commonest variant in the cases series was well within the calculated maximum alelle count in ExAC (Table 1).
+
 ```{r}
@@
 
-To assess these thresholds empirically, we explored the ExAC allele frequency spectrum of `r myVariantsN` distinct autosomal variants identified in 6179 recently published HCM cases referred for diagnostic sequencing, and individually assessed and reported according to international guidelines`r citep(c("10.1038/gim.2014.205","10.1038/gim.2016.90"))`.
+To assess the HCM thresholds empirically, we explored the ExAC allele frequency spectrum of `r myVariantsN` distinct autosomal variants identified in 6179 recently published HCM cases referred for diagnostic sequencing, and individually assessed and reported according to international guidelines`r citep(c("10.1038/gim.2014.205","10.1038/gim.2016.90"))`.
 `r myPassPath`/`r myAllPath` (`r signif(100*myPassPath/myAllPath,3)`%) variants reported as ‘Pathogenic’ or ‘Likely Pathogenic’ fell below our threshold (Figure 1), including all variants with a clear excess in cases.
@@
 
+After curating variants above our calculated HCM threshold, the false discovery rate was 0/477 (0.000; 95%CI 0.000-0.008) and 1/559 (0.002; 95%CI 0.000-0.010) for the published HCM cohort and ClinVar data respectively.
+
 #### *Extending this approach to other disorders*
 
-This approach can be readily applied in diseases where large case series are available to assess the genetic and allelic architecture, such as the inherited cardiac conditions displayed in Table 1.
 In the absence of large case series, we must estimate the genetic architecture parameters by extrapolating from similar disorders and/or variant databases.
@@
 
+### Data availability
+
+All data required to reproduce these analyses is available at <https://github.com/ImperialCardioGenetics/frequencyFilter>.
+The manuscript was compiled in R, and source code for the analysis, figures and manuscript, are available at the same location.
+Curated variant interpretations are deposited in ClinVar *Accession & DOI to be added*.
+ExAC annotations are available at <https://exac.broadinstitute.org>.
+Our allele frequency calculator app is located at <https://jamesware.shinyapps.io/alleleFrequencyApp/>, with source code available at <http://github.com/jamesware/alleleFrequencyApp>.
+
 ## DISCUSSION
@@
 Thirdly, while we believe that ExAC is depleted of severe childhood inherited conditions, and not enriched for cardiomyopathies, it could be enriched relative to the general population for some conditions, including Mendelian forms of common diseases such as diabetes or coronary disease that have been studied in contributing cohorts.
-Where this is possible, the maximum credible population allele frequency can be simply computed based on the estimated disease prevalence in the ExAC cohort, rather than the population prevalence.
+Where this is possible, the maximum credible population allele frequency should be derived based on the estimated disease prevalence in the ExAC cohort, rather than the population prevalence.
 Finally, although the resulting allele frequency thresholds are more stringent than those previously used, they are likely to still be very lenient for many applications.

### 2016-10-10 manuscript/manuscript.Rmd
discussion - limitation wrt understanding genetic architecture

@@
 In particular, we must be wary of extrapolating to or from less-well characterised populations that could harbour population-specific founder mutations.
-It is critical to define the genetic architecture in the population under study.
-We acknowledge that there is no substitute for knowning the genetic architecture of the condition of interest, and for some conditions this is not possible.
-It is the nature of any framework to rely on underlying assumptions, and we believe our method makes these explicit.
-If a parameter is particularly unclear, it is possible to define a range of architectures that fit with the observed variant.
+Unfortunately, there is no substitute for understanding the genetic architecture of the disease of interest, but when considering a specific variant that has been observed at low frequency in a given reference population, our framework and accompanying web tool provide a transparent approach to define the range of disease architectures that would be compatible with the observed data.
+For example, we may not be able to say definitively that a variant is benign, but we might be confident that it could only be causal if the prevalence of the disease is higher than published estimates, or if its penetrance is below a specified value.
+
 Secondly, it is often difficult to obtain accurate penetrance information for reported variants, and it is also difficult to know what degree of penetrance to expect or assume for newly discovered pathogenic variants (see **Supplementary Methods** for alternative approaches).

### 2016-10-11 manuscript/manuscript.Rmd
Wording changes

--- 2016-10-07 manuscript/manuscript.Rmd
@@
 We demonstrate that allele frequency cutoffs well below 0.1% are justified for a variety of human disease phenotypes and that such filters can remove an additional two-thirds of variants from consideration when compared to traditionally lenient frequency cutoffs, without discarding true pathogenic variants.
-We present pre-computed allele frequency filtering values for all variants in the Exome Aggregation Consortium database, for comparison with user-defined thresholds, which are now available through the ExAC data browser and for download, to assist others in applying our framework.
+We present pre-computed allele frequency filtering values for all variants in the Exome Aggregation Consortium database, for comparison with user-defined disease-specific thresholds, which are now available through the ExAC data browser and for download, to assist others in applying our framework.
 
@@
 Disease prevalence estimates were obtained from the literature and taken as the highest value reported.
-Variant penetrance of 0.5 was used for all analyses, as penetrance estimates for individual variants are not widely available.
-This corresponds to the reported penetrance of the HCM variant used to illustrate our approach `r citep("10.1161/CIRCRESAHA.109.216291")` and is the minimum found when researching other variants/disorders.
-When the population frequency of a variant is known, we can use this to estimate prevalence `r citep("10.1126/scitranslmed.aad5169")`.
+For cardiovascular diseases variants were modelled with a pentrance of 0.5, corresponding to the reported penetrance of the HCM variant used to illustrate our approach `r citep("10.1161/CIRCRESAHA.109.216291")` and the minimum found accross a range of variants/disorders.
 
@@
 #### *Application and validation in hypertrophic cardiomyopathy*
+#### *Defining maximum credible population allele frequency*
 
@@
 Our maximum expected population allele frequency for this allele, assuming 50% penetrance as previously reported `r citep("10.1161/CIRCRESAHA.109.216291")`, is 1/500 x 1/2 (dividing prevalence per individual by the number of chromosomes per individual) x 0.02 x 1/0.5 = 4.0x10^-5^, which we take as the maximum credible population AF for any causative variant for HCM (Table 1).
+#### *Controlling for sample variation*
 To apply this threshold while remaining robust to chance variation in observed allele counts, we ask how many times a variant with population allele frequency of 4.0x10^-5^ can be observed in a random population sample of a given size.
@@
 
-We computed the maximum tolerated allele count for four additional cardiac disorders with available cases series.
-In all cases the commonest variant in the cases series was well within the calculated maximum alelle count in ExAC (Table 1).
+#### *Assessing the acuracy of our approach*
+For all diseases with cases series that permitted us to define the genetic arcitecture, the commonest variant in the cases series was well within the calculated maximum alelle count in ExAC (Table 1).
 
@@
 `r myPassPath`/`r myAllPath` (`r signif(100*myPassPath/myAllPath,3)`%) variants reported as ‘Pathogenic’ or ‘Likely Pathogenic’ fell below our threshold (Figure 1), including all variants with a clear excess in cases.
-`r myAbsentPath` of these variants are absent from ExAC.
 The `r myFailPath` variants historically classified as ‘Likely Pathogenic’, but prevalent in ExAC in this analysis, were reassessed using contemporary ACMG criteria: there was no strong evidence in support of pathogenicity, and they were reclassified in light of these findings (Table S1).
@@
 43 variants were insufficiently rare in at least one ExAC population, and were therefore re-curated.
-42 of these had no segregation or functional data sufficient to demonstrate pathogenicity in the heterozygous state, and we would classify as VUS at most.
+42 of these had no segregation or functional data sufficient to demonstrate pathogenicity in the heterozygous state, and would be classified by the ACMG framework as VUS at most.
 The remaining variant (MYBPC3:c.3330+5G>C) had convincing evidence of pathogenicity, though with uncertain penetrance (see **Supplementary Methods**), and was observed twice in the African/African American ExAC population.
@@
 
-After curating variants above our calculated HCM threshold, the false discovery rate was 0/477 (0.000; 95%CI 0.000-0.008) and 1/559 (0.002; 95%CI 0.000-0.010) for the published HCM cohort and ClinVar data respectively.
+After curating variants above our calculated HCM threshold, the false positive rate was 0/477 (0.000; 95%CI 0.000-0.008) and 1/559 (0.002; 95%CI 0.000-0.010) for the published HCM cohort and ClinVar data respectively.
 
@@
 Intuitively, if one penetrant recessive variant were to be responsible for all PCD cases, it could have a maximum population frequency of $\sqrt(1/10000)$.
-The maximum frequency of a recessive disease-causing variant in the population can be more completely defined as:
-*max credible allele frequency* = $\sqrt(prevalence) \times$ *maximum allelic contribution* $\times \sqrt(maximum genetic contribution) \times 1/\sqrt(penetrance)$
-
-where *maximum genetic contribution* represents the proportion of all cases that are attributable to the gene under evaluation, and *maximum allelic contribution* represents the proportion of cases attributable to that gene that are attributable to an individual variant (full derivation can be found in **Methods**).
-
-We can refine our evaluation of PCD by estimating the maximum genetic and allelic contribution.
+We can refine our evaluation of PCD by estimating the maximum genetic and allelic contribution (see *Methods*).
 Across previously published cohorts of PCD cases `r citep(c("10.1164/rccm.200603-370OC","10.1164/rccm.200601-084OC","10.1038/ng.2277"))`, *DNAI1* IVS1+2\_3insT was the most common variant with a total of 17/358 alleles (4.7% 95CI 2.5-7.0%).
@@
 For each ExAC variant, we defined a "filtering allele frequency" that represents the threshold disease-specific "maximum credible allele frequency" at or below which the disease could not plausibly be caused by that variant.
-A variant with a filtering allele frequency≥the user-defined maximum credible allele frequency for the disease under consideration should be filtered, while a variant with a filtering allele frequency below the maximum credible remains a candidate (Figure 3).
-The filtering allele frequency is not desease speific and hence can be applied to any disease of interest.
+A variant with a filtering allele frequency≥the maximum credible allele frequency for the disease under consideration should be filtered, while a variant with a filtering allele frequency below the maximum credible remains a candidate.
+The filtering allele frequency is not desease speific and hence can be applied to any disease of interest by comparing with a user-defined disease-specific maximum credible allele frequency (Figure 3).
 This value has been pre-computed for all variants in ExAC (see **Methods** and **Supplementary Methods**), and is available via the ExAC VCF and browser (http://exac.broadinstitute.org).
@@
 The odds ratio for disease-association increases markedly at very low allele frequencies (Figure 2b) demonstrating that increasing the stringency of a frequency filter improves the information content of a genetic result.
-
-### Data availability
-
-All data required to reproduce these analyses is available at <https://github.com/ImperialCardioGenetics/frequencyFilter>.
-The manuscript was compiled in R, and source code for the analysis, figures and manuscript, are available at the same location.
-Curated variant interpretations are deposited in ClinVar *Accession & DOI to be added*.
-ExAC annotations are available at <https://exac.broadinstitute.org>.
-Our allele frequency calculator app is located at <https://jamesware.shinyapps.io/alleleFrequencyApp/>, with source code available at <http://github.com/jamesware/alleleFrequencyApp>.
 
@@
 We recognise several limitations of our approach.
-First, the approach is limited by our understanding of the prevalence and genetic architecture of the disease in question: this characterisation will vary for different diseases and in different populations, though we illustrate approaches for estimation and extrapolation of parameters.
+First, we are limited by our understanding of the prevalence and genetic architecture of the disease in question: this characterisation will vary for different diseases and in different populations, though we illustrate approaches for estimation and extrapolation of parameters.
 In particular, we must be wary of extrapolating to or from less-well characterised populations that could harbour population-specific founder mutations.
-It is critical to define the genetic architecture in the population under study.
-We acknowledge that there is no substitute for knowning the genetic architecture of the condition of interest, and for some conditions this is not possible.
-It is the nature of any framework to rely on underlying assumptions, and we believe our method makes these explicit.
-If a parameter is particularly unclear, it is possible to define a range of architectures that fit with the observed variant.
-Secondly, it is often difficult to obtain accurate penetrance information for reported variants, and it is also difficult to know what degree of penetrance to expect or assume for newly discovered pathogenic variants (see **Supplementary Methods** for alternative approaches).
+Unfortunately, there is no substitute for understanding the genetic architecture of the disease of interest, but when considering a specific variant that has been observed at low frequency in a given reference population, our framework and accompanying web tool provide a transparent approach to define the range of disease architectures that are compatible with the observed data.
+For example, we may not be able to say definitively that a variant is benign, but we might be confident that it could only be causal if the prevalence of the disease is higher than published estimates, or its penetrance is below a specified value.
+
+Secondly, it is often difficult to obtain accurate penetrance information for reported variants, and it is also difficult to know what degree of penetrance to expect or assume for newly discovered pathogenic variants.
+Although we would argue that variants with low penetrance have questionable diagnostic utility, our calculator app allows a user to define a range of compatible penetrance for a given allele frequency (see **Supplementary Methods**).
+Furthermore, methods to estimate variant penetrance from large case and control cohorts have been described previously`r citep("10.1126/scitranslmed.aad5169")`.
 
@@
 Increased phenotypic information linked to reference datasets will also reduce limitations due to uncertain disease status, and improve prevalence estimates, adding further power to our approach.
+
+## DATA AVAILABILITY
+
+All data required to reproduce these analyses is available at <https://github.com/ImperialCardioGenetics/frequencyFilter>.
+The manuscript was compiled in R, and source code for the analysis, figures and manuscript, are available at the same location.
+Curated variant interpretations are deposited in ClinVar *Accession & DOI to be added*.
+ExAC annotations are available at <https://exac.broadinstitute.org>.
+Our allele frequency calculator app is located at <https://jamesware.shinyapps.io/alleleFrequencyApp/>, with source code available at <http://github.com/jamesware/alleleFrequencyApp>.
 

### 2016-10-11 manuscript/manuscript.Rmd
Wording changes

@@
 We recognise several limitations of our approach.
-First, we are limited by our understanding of the prevalence and genetic architecture of the disease in question: this characterisation will vary for different diseases and in different populations, though we illustrate approaches for estimation and extrapolation of parameters.
+First, the approach is limited by our understanding of the prevalence and genetic architecture of the disease in question: this characterisation will vary for different diseases and in different populations, though we illustrate approaches for estimation and extrapolation of parameters.
 In particular, we must be wary of extrapolating to or from less-well characterised populations that could harbour population-specific founder mutations.
-Unfortunately, there is no substitute for understanding the genetic architecture of the disease of interest, but when considering a specific variant that has been observed at low frequency in a given reference population, our framework and accompanying web tool provide a transparent approach to define the range of disease architectures that are compatible with the observed data.
-For example, we may not be able to say definitively that a variant is benign, but we might be confident that it could only be causal if the prevalence of the disease is higher than published estimates, or its penetrance is below a specified value.
+Unfortunately, there is no substitute for understanding the genetic architecture of the disease of interest, but when considering a specific variant that has been observed at low frequency in a given reference population, our framework and accompanying web tool provide a transparent approach to define the range of disease architectures that would be compatible with the observed data.
+For example, we may not be able to say definitively that a variant is benign, but we might be confident that it could only be causal if the prevalence of the disease is higher than published estimates, or if its penetrance is below a specified value.
 
@@
 Although we would argue that variants with low penetrance have questionable diagnostic utility, our calculator app allows a user to define a range of compatible penetrance for a given allele frequency (see **Supplementary Methods**).
-Furthermore, methods to estimate variant penetrance from large case and control cohorts have been described previously`r citep("10.1126/scitranslmed.aad5169")`.
+Furthermore, methods to estimate variant penetrance from large case and control cohorts have been described previously `r citep("10.1126/scitranslmed.aad5169")`.
 

### 2016-11-01 manuscript/SupplementaryMaterial.Rmd
Removed more words
More reformatting
Edits from JW

@@
 
-## Supplementary Table S1
-
-Variants previously reported as causative of HCM either in ClinVar, or in a clinical series of 6179 HCM cases, but that were observed in ExAC above the maximum tolerated allele count for HCM (AC>9 globally) were manually curated according to the ACMG guidelines for interpretation and classification on sequence variants.
-*These variants were previously categorised as Likely Pathogenic in the LMM and Oxford HCM cohorts.
-
-```{r}
-read.delim("../data-raw/SupplementaryTable1.txt") %>% kable
-```
-
 ## Suplementary Methods
@@
 where *maximum genetic contribution* represents the proportion of all cases that are attributable to the gene under evaluation, and *maximum allelic contribution* represents the proportion of cases attributable to that gene that are attributable to an individual variant.
-The maximum frequency of a recessive disease causing variant in the population was therefore defined as:
-*max credible AF* = $\sqrt(prevalence) \times$ *maximum allelic contribution* $\times \sqrt(maximum genetic contribution) \times 1/\sqrt(penetrance)$
+This can be rearranged to give the maximum frequency of a recessive disease causing variant in the population as:
+*max credible AF* = $\sqrt($*prevalence*$) \times$ *maximum allelic contribution* $\times \sqrt($*maximum genetic contribution*$) \times 1/\sqrt($*penetrance*$)$
 
@@
 It is often difficult to obtain accurate penetrance information for reported variants, and it is also difficult to know what degree of penetrance to expect or assume in newly discovered pathogenic variants.
-In this work we uniformly apply a value of 50% penetrance (i.e.
+In this work we uniformly apply a value of 50% penetrance for inherited cardiac conditions (i.e.
 assuming variant penetrance is no lower than 50%) equivalent to that reported for our HCM example variant, and the lowest we found reported for any of our examples or reported accross our studied disorders.
@@
 
+## Supplementary Table S1
+
+Variants previously reported as causative of HCM either in ClinVar, or in a clinical series of 6179 HCM cases, but that were observed in ExAC above the maximum tolerated allele count for HCM (AC>9 globally) were manually curated according to the ACMG guidelines for interpretation and classification on sequence variants.
+*These variants were previously categorised as Likely Pathogenic in the LMM and Oxford HCM cohorts.
+
+```{r}
+read.delim("../data-raw/SupplementaryTable1.txt") %>% kable
+```
+
 ##Supplemental references

### 2016-11-01 manuscript/manuscript.Rmd
Edits from JW

--- 2016-10-11 manuscript/manuscript.Rmd
@@
 
-Key words: Variant interpretation, Clinical Genomics, Allele frequency, Cardiomyopathy, Inherited Cardiovascular conditions, ExAC
+Key words: Variant interpretation, Clinical Genomics, Allele frequency, Cardiomyopathy, Inherited Cardiovascular Conditions, ExAC
 
@@
 A variant's low frequency in, or absence from, reference databases is recognised as a necessary, but not sufficient, criterion for variant pathogenicity `r citep(c("10.1038/gim.2015.30","10.1038/nature13127"))`.
-The recent availability of very large reference databases, such as the Exome Aggregation Consortium (ExAC) `r citep("10.1038/nature19057")` dataset, which has characterised the population allele frequencies of 10 million genomic variants through the analysis of exome sequencing data from over 60,000 humans, provides an opportunity to obtain robust frequency estimates even for rare variants, improving the theoretical power for allele frequency filtering in Mendelian variant discovery efforts.
+The recent availability of very large reference databases, such as the Exome Aggregation Consortium (ExAC) `r citep("10.1038/nature19057")` dataset, which has characterised the population allele frequencies of 10 million genomic variants through the analysis of exome sequencing data from over 60,000 humans, provides an opportunity to obtain robust frequency estimates even for rare variants, improving the theoretical power for allele frequency (AF) filtering in Mendelian variant discovery efforts.
 
-In practice, there exists considerable ambiguity around what allele frequency should be considered "too common", with the lenient values of 1% and 0.1% often invoked as conservative frequency cutoffs for recessive and dominant diseases respectively `r citep("10.1038/nrg3031")`.
+In practice, there exists considerable ambiguity around what AF should be considered "too common", with the lenient values of 1% and 0.1% often invoked as conservative frequency cutoffs for recessive and dominant diseases respectively `r citep("10.1038/nrg3031")`.
 Population genetics, however, dictates that severe disease-causing variants must be much rarer than these cutoffs, except in cases of bottlenecked populations, balancing selection, or other special circumstances `r citep(c("10.1093/molbev/msp190","10.1073/pnas.1322563111"))`.
@@
 This rule must, however, be refined to account for different inheritance modes, genetic and allelic heterogeneity, and reduced penetrance.
-In addition, for rare variants, estimation of true population allele frequency is clouded by considerable sampling variance, even in the largest samples currently available.
-These limitations have encouraged the adoption of very lenient approaches when filtering variants by allele frequency `r citep(c("10.1016/S0140-6736(14)61705-0","10.1038/ng.3304"))`, and recognition that more stringent approaches that account for disease-specific genetic architecture are urgently needed`r citep("10.1016/j.ajhg.2016.03.024")`.
+In addition, for rare variants, estimation of true population AF is clouded by considerable sampling variance, even in the largest samples currently available.
+These limitations have encouraged the adoption of very lenient approaches when filtering variants by AF `r citep(c("10.1016/S0140-6736(14)61705-0","10.1038/ng.3304"))`, and recognition that more stringent approaches that account for disease-specific genetic architecture are urgently needed`r citep("10.1016/j.ajhg.2016.03.024")`.
 
 Here we present a statistical framework for assessing whether rare variants are sufficiently rare to cause penetrant Mendelian disease, while accounting for both architecture and sampling variance in observed allele counts.
-We demonstrate that allele frequency cutoffs well below 0.1% are justified for a variety of human disease phenotypes and that such filters can remove an additional two-thirds of variants from consideration when compared to traditionally lenient frequency cutoffs, without discarding true pathogenic variants.
-We present pre-computed allele frequency filtering values for all variants in the Exome Aggregation Consortium database, for comparison with user-defined disease-specific thresholds, which are now available through the ExAC data browser and for download, to assist others in applying our framework.
+We demonstrate that AF cutoffs well below 0.1% are justified for a variety of human disease phenotypes and that such filters can remove an additional two-thirds of variants from consideration when compared to traditionally lenient frequency cutoffs, without discarding true pathogenic variants.
+We present pre-computed AF filtering values for all variants in the Exome Aggregation Consortium database, for comparison with user-defined disease-specific thresholds, which are now available through the ExAC data browser and for download, to assist others in applying this framework.
 
@@
 
+We define a two-stage approach to determine whether a variant observed in a reference sample is too common to cause a given disease.
+First, we define a maximum population AF that we believe is credible for a pathogenic variant, given the genetic architecture of the disease in question.
+Second, we determine whether the observed allele count in our reference sample is consistent with a variant having this frequency in the population from which the sample was drawn.
+
 For a penetrant dominant Mendelian allele to be disease causing, it cannot be present in the general population more frequently that the disease it causes.
 Furthermore, if the disease is genetically heterogeneous, it must not be more frequent than the proportion of cases attributable to that gene, or indeed to any single variant.
-We can therefore define the maximum credible population allele frequency (for a pathogenic allele) as:
+We can therefore define the maximum credible population AF (for a pathogenic allele) as:
 
@@
 For recessive conditions, the maximum AF is defined as:
-*max credible allele frequency* = $\sqrt(prevalence) \times$ *maximum allelic contribution* $\times \sqrt(maximum genetic contribution) \times 1/\sqrt(penetrance)$
+*maximum credible population AF* = $\sqrt($*prevalence*$) \times$ *maximum allelic contribution* $\times \sqrt($*maximum genetic contribution*$) \times 1/\sqrt($*penetrance*$)$
 
-where *maximum genetic contribution* represents the proportion of all cases that are attributable to the gene under evaluation, and *maximum allelic contribution* represents the proportion of cases attributable to that gene that are attributable to an individual variant.
-Full derivation can be found in **Supplementary Methods**.
+where *maximum genetic contribution* represents the proportion of all cases that are attributable to the gene under evaluation, and *maximum allelic contribution* represents the proportion of cases attributable to that gene that are attributable to an individual variant (see **Supplementary Methods** for full derivation).
 
 Disease prevalence estimates were obtained from the literature and taken as the highest value reported.
-For cardiovascular diseases variants were modelled with a pentrance of 0.5, corresponding to the reported penetrance of the HCM variant used to illustrate our approach `r citep("10.1161/CIRCRESAHA.109.216291")` and the minimum found accross a range of variants/disorders.
+Cardiovascular diseases variants were modelled with a pentrance of 0.5, corresponding to the reported penetrance of the HCM variant used to illustrate our approach `r citep("10.1161/CIRCRESAHA.109.216291")` and the minimum found accross a range of variants/disorders.
 
-We do not know the true population allele frequency of any variant, having only an observed allele frequency in a finite population sample.
+We do not know the true population AF of any variant, having only an observed AF in a finite population sample.
 Moreover, confidence intervals around this observed frequency are problematic to estimate given our incomplete knowledge of the frequency spectrum of rare variants, which appears to be skewed towards very rare variants.
@@
 
-If we turn the problem around, and begin instead from allele frequency, specifying a maximum _true_ allele frequency value we are willing to consider in the population (using the equation above), then we can estimate the probability distribution for allele counts in a given sample size.
+To address this, we begin by specifying a maximum _true_ AF value we are willing to consider in the population (using the equation above), from which we can estimate the probability distribution for allele counts in a given sample size.
 This follows a binomial distribution, and can be satisfactorily approximated with a Poisson distribution (see **Supplementary Methods**).
-This allows us to set an upper limit on the number of alleles in a sample that is consistent with a given population frequency.
+This allows us to set an upper limit on the number of alleles in a sample that is consistent with a given underlying population frequency.
+For example, a variant with a true population AF of 0.0001 would be expected to occur in a sample of 100,000 alleles ≤15 times with a probability of 0.95.
 
-The maximum tolerated allele count (AC) was therefore computed as the AC at the upper bound of the one-tailed 95% confidence interval (95%CI AC) of a Poisson distribution, for the established maximum credible AF, given the observed allele number (AN).
-In R: `max_ac = qpois(quantile_limit,an*af)`, where `max_ac` is the 95%CI AC, `quantile_limit` is 0.95 (for a one-sided 95%CI), `an` is the observed allele number, and `af` is the maximum credible population AF.
+We therefore computed a maximum tolerated allele count (AC) as the AC at the upper bound of the one-tailed 95% confidence interval (95%CI AC) of a Poisson distribution, for the specified maximum credible AF, given the sample size (observed allele number, AN).
 
@@
 
-In order to pre-compute `af_filter` values for all of ExAC (verson 0.3.1), we apply a two-step approach to the AC and AN values for each of the five major continental populations, and take the highest result from any population (more explanation in **Supplementary Methods**).
+We can reverse this process to determine the maximum true population AF that is consistent with a particular observed sample AC, and we applied this to the ExAC dataset (version 0.3.1).
+In order to pre-compute `af_filter` values for all variants in ExAC, we apply a two-step approach to the AC and AN values for each of the five major continental populations, and take the highest result from any population (more explanation in **Supplementary Methods**).
 
@@
 
-To simulate Mendelian variant discovery, we randomly selected 100 individuals from each of five major continental populations and filtered their exomes against filtering allele frequencies derived from the remaining 60,206 ExAC individuals.
+To simulate Mendelian variant discovery, we randomly selected 100 individuals from each of five major continental populations and filtered their exomes against filtering AFs derived from the remaining 60,206 ExAC individuals.
 The subset of individuals was the same as that previously reported `r citep("10.1038/nature19057")`.
@@
 
-Pathogenic and non-conflicted variants from XML and TXT releases of ClinVar (July 9, 2015 release) were extracted using a Python implementation of vt normalize `r citep("10.1093/bioinformatics/btv112")`, as described previously `r citep("10.1038/nature19057")`.
+Pathogenic and non-conflicted variants were extracted from ClinVar (July 9, 2015 release) as described previously `r citep("10.1038/nature19057")`.
 ExAC counts were determined by matching on chromosome, position, reference, and alternate alleles.
@@
 #### *Application and validation in hypertrophic cardiomyopathy*
-#### *Defining maximum credible population allele frequency*
+#### *Defining maximum credible population AF*
 
-We illustrate our general approach using the dominant cardiac disorder hypertrophic cardiomyopathy (HCM), which has an estimated prevalence of 1 in 500 in the general population `r citep("10.1038/gim.2014.205")`.
-As there have been previous large-scale genetic studies of HCM, with series of up to 6,179 individuals `r citep(c("10.1038/gim.2014.205","10.1038/gim.2016.90"))`, we can make the assumption that no newly identified variant will be more frequent in cases that those identified to date (at least for well-studied ancestries).
-This allows us to define the maximum allelic contribution of any single variant to the disorder.
-In these large case series, the largest proportion of cases is attributable to the missense variant _MYBPC3_ c.1504C>T (p.Arg502Trp), found in 104/6179 HCM cases (1.7%; 95%CI 1.4-2.0%) `r citep(c("10.1038/gim.2014.205","10.1038/gim.2016.90"))`.
+We illustrate our generalisable approach using the dominant cardiac disorder hypertrophic cardiomyopathy (HCM), which has an estimated prevalence of 1 in 500 in the general population `r citep("10.1038/gim.2014.205")`.
+As there have been previous large-scale genetic studies of HCM, with series of up to 6,179 individuals `r citep(c("10.1038/gim.2014.205","10.1038/gim.2016.90"))`, we can assume that no newly identified variant will be more frequent in cases that those identified to date (at least for well-studied ancestries), allowing us to define the maximum contribution of any single variant to the disorder.
+In these series, the largest proportion of cases is attributable to the missense variant _MYBPC3_ c.1504C>T (p.Arg502Trp), found in 104/6179 HCM cases (1.7%; 95%CI 1.4-2.0%) `r citep(c("10.1038/gim.2014.205","10.1038/gim.2016.90"))`.
 We therefore take the upper bound of this proportion (0.02) as an estimate of the maximum allelic contribution in HCM (Table 1).
-Our maximum expected population allele frequency for this allele, assuming 50% penetrance as previously reported `r citep("10.1161/CIRCRESAHA.109.216291")`, is 1/500 x 1/2 (dividing prevalence per individual by the number of chromosomes per individual) x 0.02 x 1/0.5 = 4.0x10^-5^, which we take as the maximum credible population AF for any causative variant for HCM (Table 1).
+Our maximum expected population AF for this allele, assuming penetrance 0.5 as previously reported `r citep("10.1161/CIRCRESAHA.109.216291")`, is 1/500 x 1/2 (dividing prevalence per individual by the number of chromosomes per individual) x 0.02 x 1/0.5 = 4.0x10^-5^, which we take as the maximum credible population AF for any causative variant for HCM (Table 1).
+
 #### *Controlling for sample variation*
-To apply this threshold while remaining robust to chance variation in observed allele counts, we ask how many times a variant with population allele frequency of 4.0x10^-5^ can be observed in a random population sample of a given size.
-For a 5% error rate we take the 95th percentile of a poisson distribution with λ = expected allele count, which is given by: sample size (chromosomes) &times; expected population allele frequency (**Methods**).
+
+To apply this threshold while remaining robust to chance variation in observed allele counts, we ask how many times a variant with population AF of 4.0x10^-5^ can be observed in a random population sample of a given size.
+For a 5% error rate we take the 95th percentile of a poisson distribution with λ = expected allele count, which is given by: sample size (chromosomes) &times; expected population AF (**Methods**).
 For HCM this gives us a maximum tolerated allele count of 9, assuming 50% penetrance (or 5 for fully penetrant alleles), for variants genotyped in the full ExAC cohort (sample size=121,412 chromosomes).
@@
 
-To facilitate these calculations, we have produced an online calculator (<https://jamesware.shinyapps.io/alleleFrequencyApp/>) that will compute maximum credible population allele frequency and maximum sample allele count for a user-specified genetic architecture, and conversely allow users to dynamically explore what genetic architecture(s) might be most compatible with an observed variant having a causal role in disease.
+To facilitate these calculations, we have produced an online calculator (<https://jamesware.shinyapps.io/alleleFrequencyApp/>) that will compute maximum credible population AF and maximum sample allele count for a user-specified genetic architecture, and conversely allow users to dynamically explore what genetic architecture(s) might be most compatible with an observed variant having a causal role in disease.
 
 #### *Assessing the acuracy of our approach*
+
 For all diseases with cases series that permitted us to define the genetic arcitecture, the commonest variant in the cases series was well within the calculated maximum alelle count in ExAC (Table 1).
@@
 
-To assess the HCM thresholds empirically, we explored the ExAC allele frequency spectrum of `r myVariantsN` distinct autosomal variants identified in 6179 recently published HCM cases referred for diagnostic sequencing, and individually assessed and reported according to international guidelines`r citep(c("10.1038/gim.2014.205","10.1038/gim.2016.90"))`.
+To assess the HCM thresholds empirically, we explored the ExAC AF spectrum of `r myVariantsN` distinct autosomal variants, identified in 6179 published HCM cases referred for diagnostic sequencing, and individually assessed and clinically reported according to international guidelines`r citep(c("10.1038/gim.2014.205","10.1038/gim.2016.90"))`.
 `r myPassPath`/`r myAllPath` (`r signif(100*myPassPath/myAllPath,3)`%) variants reported as ‘Pathogenic’ or ‘Likely Pathogenic’ fell below our threshold (Figure 1), including all variants with a clear excess in cases.
 The `r myFailPath` variants historically classified as ‘Likely Pathogenic’, but prevalent in ExAC in this analysis, were reassessed using contemporary ACMG criteria: there was no strong evidence in support of pathogenicity, and they were reclassified in light of these findings (Table S1).
-This analysis identifies `r myFailVus`/`r myAllVus` (`r signif(100*myFailVus/myAllVus,3)`%) VUS that are unlikely to be truly causative for HCM.
+This analysis identifies `r myFailVus`/`r myAllVus` (`r signif(100*myFailVus/myAllVus,3)`%) VUS that are very unlikely to be causative for HCM.
 
-The above analysis applied a single global allele count limit of 9 for HCM, however, as allele frequencies differ between populations, filtering based on frequencies in individual populations may provide greater power `r citep("10.1038/nature19057")`.
+The above analysis applied a single global allele count limit of 9 for HCM, however, as AFs differ between populations, filtering based on frequencies in individual populations may provide greater power `r citep("10.1038/nature19057")`.
 For example, a variant relatively common in any one population is unlikely pathogenic, even if rare in other populations, provided the disease prevalence and architecture is consistent across populations.
-We therefore compute a maximum tolerated AC for each distinct sub-population of our reference sample, and filter based on the highest allele frequency observed in any major continental population (see **Methods** and **Suplementary Note 3**).
+We therefore compute a maximum tolerated AC for each distinct sub-population of our reference sample, and filter based on the highest AF observed in any major continental population (see **Methods** and **Suplementary Note 3**).
 
@@
 43 variants were insufficiently rare in at least one ExAC population, and were therefore re-curated.
-42 of these had no segregation or functional data sufficient to demonstrate pathogenicity in the heterozygous state, and would be classified by the ACMG framework as VUS at most.
+42 of these had no segregation or functional data sufficient to demonstrate pathogenicity in the heterozygous state, and would be classified by the contemporary ACMG framework as VUS at most.
 The remaining variant (MYBPC3:c.3330+5G>C) had convincing evidence of pathogenicity, though with uncertain penetrance (see **Supplementary Methods**), and was observed twice in the African/African American ExAC population.
@@
 
-In the absence of large case series, we must estimate the genetic architecture parameters by extrapolating from similar disorders and/or variant databases.
+This framework relies on estimation of the genetic architecture of a condition, which may not be well described.
+For diseases where large case series are absent, we can estimate the genetic architecture parameters by extrapolating from similar disorders and/or variant databases.
 
-Where disease-specific variant databases exist, we can use these to help estimate the maximum allelic contribution in lieu of individual case series.
+Where disease-specific variant databases exist, we can use these to estimate the maximum allelic contribution.
 For example, Marfan syndrome is a rare connective tissue disorder caused by variants in the _FBN1_ gene.
@@
 Where no mutation database exists, we can use what is known about similar disorders to estimate the maximum allelic contribution.
-For the cardiac conditions with large cases series in Table 1, the maximum proportion of cases attributable to any one variant is 6.7% (95CI 4.1-9.2%; PKP2:c.2146-1G>C found in 24/361 ARVC cases `r citep("10.1038/gim.2016.90")`).
-We therefore take the upper bound of this confidence interval (rounded up to 0.1) as an estimate of the maximum allelic contribution for other genetically heterogeneous cardiac conditions, unless we can find disease-specific evidence to alter it.
+For the better-characterised cardiac conditions in Table 1, the maximum proportion of cases attributable to any one variant is 6.7% (95CI 4.1-9.2%; PKP2:c.2146-1G>C found in 24/361 ARVC cases `r citep("10.1038/gim.2016.90")`).
+We therefore propose the upper bound of this confidence interval (rounded up to 0.1) as a resonable estimate of the maximum allelic contribution for other genetically heterogeneous cardiac conditions, unless there is disease-specific evidence to alter it.
 For Noonan syndrome and Catecholaminergic Polymorphic Ventricular Tachycardia (CPVT - an inherited cardiac arrhythmia syndrome) with prevalences of 1 in 1000 `r citep("10.1016/S0140-6736(12)61023-X")` and 1 in 10,000 `r citep("10.1038/ejhg.2013.55")` respectively, this translates to maximum population frequencies of 5x10^-5^ and 5x10^-6^ and maximum tolerated ExAC allele counts of 10 and 2 (Table 2).
@@
 *DNAI1* IVS1+2\_3insT is itself present at 56/121108 ExAC alleles (45/66636 non-Finnish European alleles).
-A single variant reported to cause PCD in ClinVar occurs in ExAC with AC > 332 (_NME8_ NM\_016616.4:c.271-27C>T; AC=2306/120984): our model therefore indicates that this variant frequency is too common to be disease-causing, and consistent with this we note that it meets none of the current ACMG criteria for assertions of pathogenicity, and have reclassified it as VUS (see **Supplementary Methods**).
+A single variant reported to cause PCD in ClinVar occurs in ExAC with AC > 332 (_NME8_ NM\_016616.4:c.271-27C>T; AC=2306/120984): our model therefore indicates that this variant frequency is too common to be disease-causing, and consistent with this we note that it meets none of the current ACMG criteria for assertions of pathogenicity, and would reclassify it as VUS (see **Supplementary Methods**).
 
@@
 
-For each ExAC variant, we defined a "filtering allele frequency" that represents the threshold disease-specific "maximum credible allele frequency" at or below which the disease could not plausibly be caused by that variant.
-A variant with a filtering allele frequency≥the maximum credible allele frequency for the disease under consideration should be filtered, while a variant with a filtering allele frequency below the maximum credible remains a candidate.
-The filtering allele frequency is not desease speific and hence can be applied to any disease of interest by comparing with a user-defined disease-specific maximum credible allele frequency (Figure 3).
+For each ExAC variant, we defined a "filtering AF" that represents the threshold disease-specific "maximum credible AF" at or below which the disease could not plausibly be caused by that variant.
+A variant with a filtering AF ≥ the maximum credible AF for the disease under consideration should be filtered, while a variant with a filtering AF below the maximum credible remains a candidate.
+This filtering AF is not disease specific: it can be applied to any disease of interest by comparing with a user-defined disease-specific maximum credible AF (Figure 3).
 This value has been pre-computed for all variants in ExAC (see **Methods** and **Supplementary Methods**), and is available via the ExAC VCF and browser (http://exac.broadinstitute.org).
 
-To assess the efficiency of our approach, we calculated the filtering allele frequency on 60,206 exomes from ExAC and applied these filters to a simulated dominant Mendelian variant discovery analysis on the remaining 500 exomes (see **Methods**).
-Filtering at allele frequencies lower than 0.1% substantially reduces the number of predicted protein-altering variants in consideration, with the mean number of variants per exome falling from 176 to 63 at cutoffs of 0.1% and 0.0001% respectively (Figure 2a).
-Additionally, we compared the prevalence of variants in HCM genes in cases and controls across the allele frequency spectrum, and computed disease odds ratios for different frequency bins.
-The odds ratio for disease-association increases markedly at very low allele frequencies (Figure 2b) demonstrating that increasing the stringency of a frequency filter improves the information content of a genetic result.
+To assess the efficiency of our approach, we calculated the filtering AF on 60,206 exomes from ExAC and applied these filters to a simulated dominant Mendelian variant discovery analysis on the remaining 500 exomes (see **Methods**).
+Filtering at AFs lower than 0.1% substantially reduces the number of predicted protein-altering variants in consideration, with the mean number of variants per exome falling from 176 to 63 at cutoffs of 0.1% and 0.0001% respectively (Figure 2a).
+Additionally, we compared the prevalence of variants in HCM genes in cases and controls across the AF spectrum, and computed disease odds ratios for different frequency bins.
+The odds ratio for disease-association increases markedly at very low AFs (Figure 2b) demonstrating that increasing the stringency of a frequency filter improves the information content of a genetic result.
 
@@
 
-To assist others in applying our framework, we have precomputed a 'filtering allele frequency' for all variants across the ExAC dataset.
-This is defined such that if the filtering allele frequency of a variant is at or above the user-defined "maximum credible population allele frequency" for the disease in question, then that variant is not a credible candidate (in other words, for any population allele frequency below the threshold value, the probability of the observed allele count in the ExAC sample is <0.05).
-Once a user has determined their "maximum credible population allele frequency", they may remove from consideration ExAC variants for which the filtering allele frequency is greater than or equal to than the chosen value.
+To assist others in applying our framework, we have precomputed a 'filtering AF' for all variants across the ExAC dataset.
+This is defined such that if the filtering AF of a variant is at or above the user-defined "maximum credible population AF" for the disease in question, then that variant is not a credible candidate (in other words, for any population AF below the threshold value, the probability of the observed allele count in the ExAC sample is <0.05).
+Once a user has determined their "maximum credible population AF", they may remove from consideration ExAC variants for which the filtering AF is greater than or equal to than the chosen value.
 
@@
 In particular, we must be wary of extrapolating to or from less-well characterised populations that could harbour population-specific founder mutations.
-Unfortunately, there is no substitute for understanding the genetic architecture of the disease of interest, but when considering a specific variant that has been observed at low frequency in a given reference population, our framework and accompanying web tool provide a transparent approach to define the range of disease architectures that are compatible with the observed data.
+While incomplete knowledge of the genetic architecture of a disease of interest will limit this or any approach to evaluate a specific variant that has been observed at low frequency in a reference population, our framework and accompanying web tool do at least transparently define the range of disease architectures that are compatible with the observed data.
 For example, we may not be able to say definitively that a variant is benign, but we might be confident that it could only be causal if the prevalence of the disease is higher than published estimates, or its penetrance is below a specified value.
@@
 Secondly, it is often difficult to obtain accurate penetrance information for reported variants, and it is also difficult to know what degree of penetrance to expect or assume for newly discovered pathogenic variants.
-Although we would argue that variants with low penetrance have questionable diagnostic utility, our calculator app allows a user to define a range of compatible penetrance for a given allele frequency (see **Supplementary Methods**).
-Furthermore, methods to estimate variant penetrance from large case and control cohorts have been described previously`r citep("10.1126/scitranslmed.aad5169")`.
+Although we would argue that variants with low penetrance have questionable diagnostic utility, our calculator app allows a user to define a range of compatible penetrance for a given AF (see **Supplementary Methods**), and implements methods to estimate variant penetrance from prevalence data in case and control cohorts as previously described `r citep("10.1126/scitranslmed.aad5169")`.
 
 Thirdly, while we believe that ExAC is depleted of severe childhood inherited conditions, and not enriched for cardiomyopathies, it could be enriched relative to the general population for some conditions, including Mendelian forms of common diseases such as diabetes or coronary disease that have been studied in contributing cohorts.
-Where this is possible, the maximum credible population allele frequency should be derived based on the estimated disease prevalence in the ExAC cohort, rather than the population prevalence.
-Finally, although the resulting allele frequency thresholds are more stringent than those previously used, they are likely to still be very lenient for many applications.
+Where this is possible, the maximum credible population AF should be derived based on the estimated disease prevalence in the ExAC cohort, rather than the population prevalence.
+
+Finally, although the resulting AF thresholds are more stringent than those previously used, they are likely to still be very lenient for many applications.
 For instance, we base our calculation on the most prevalent known pathogenic variant from a disease cohort.
@@
 Increases in both the ancestral diversity and size of reference datasets will bring additional power to our method over time.
-We have avoided filtering on variants observed only once, because a single observation provides little information about true allele frequency (see **Supplementary Methods**).
+We have avoided filtering on variants observed only once, because a single observation provides little information about true AF (see **Supplementary Methods**).
 A ten-fold increase in sample size, resulting from projects such as the US Precision Medicine Initiative, will separate vanishingly rare variants from those whose frequency really is ~1 in 100,000.
@@
 ExAC annotations are available at <https://exac.broadinstitute.org>.
-Our allele frequency calculator app is located at <https://jamesware.shinyapps.io/alleleFrequencyApp/>, with source code available at <http://github.com/jamesware/alleleFrequencyApp>.
+Our AF calculator app is located at <https://jamesware.shinyapps.io/alleleFrequencyApp/>, with source code available at <http://github.com/jamesware/alleleFrequencyApp>.
 
@@
 
-This work was supported by the Wellcome Trust (107469/Z/15/Z), the Medical Research Council (UK), the NIHR Biomedical Research Unit in Cardiovascular Disease at Royal Brompton & Harefield NHS Foundation Trust and Imperial College London, the Fondation Leducq (11 CVD-01), a Health Innovation Challenge Fund (HICF-R6-373) award from the Wellcome Trust and Department of Health, UK, and by the National Institute of Diabetes and Digestive and Kidney Diseases and the National Institute of General Medical Sciences of the NIH (awards U54DK105566 and R01GM104371).
+This work was supported by the Wellcome Trust (107469/Z/15/Z), the Medical Research Council (UK), the NIHR Biomedical Research Unit in Cardiovascular Disease at Royal Brompton & Harefield NHS Foundation Trust and Imperial College London, the Fondation Leducq (11 CVD-01), a Health Innovation Challenge Fund award from the Wellcome Trust and Department of Health, UK (HICF-R6-373), and by the National Institute of Diabetes and Digestive and Kidney Diseases and the National Institute of General Medical Sciences of the NIH (awards U54DK105566 and R01GM104371).
 EVM is supported by the National Institutes of Health under a Ruth L.
@@
 - Treatment of singletons and other populations
-- Description of the filtering allele frequency
+- Description of the filtering AF
 
@@
 > The clinical utility of stringent allele frequency thresholds.
-(a) The number of predicted protein-altering variants (definition in Methods) per exome as a function of the frequency filter applied.
+(a) The number of predicted protein-altering variants (definition in Methods) per exome as a function of the allele frequency filter applied.
 A one-tailed 95% confidence interval is used, meaning that variants were removed from consideration if their AC would fall within the top 5% of the Poisson probability distribution for the user's maximum credible AF (x axis).
 (b) The odds ratio for HCM disease-association against allele frequency.
-The prevalence of variants in HCM-associated genes (_MYH7_, _MYBPC3_ and other sarcomeric  (_TNNT2_, _TNNI3_, _MYL2_, _MYL3_, _TPM1_ and _ACTC1_, analysed collectively) in 322 HCM cases and 60,706 ExAC controls were compared for a range of allele frequency bins, and an odds ratio computed (see **Methods**).
+The prevalence of variants in HCM-associated genes (_MYH7_, _MYBPC3_ and other sarcomeric  (_TNNT2_, _TNNI3_, _MYL2_, _MYL3_, _TPM1_ and _ACTC1_, analysed collectively)) in 322 HCM cases and 60,706 ExAC controls were compared for a range of allele frequency bins, and an odds ratio computed (see **Methods**).
 Data for each bin is plotted at the upper allele frequency cutoff.
@@
 
-> A flow diagram of our approach, applied to a dominant condition.
+> A flow diagram of our approach, applied to a dominant condition, and using ExAC as our reference sample.
 First, a disease-level maximum credible population allele frequency is calculated, based on disease prevalence, heterogeneity and penetrance.
-This is then compared against the pre-calculated variant filtering allele frequency to determine if a variant is compatible with disease.
+To evaluate a specific variant, we determine whether the observed variant allele count is compatible with disease by comparing this maximum credible population AF against the (pre-calculated) filtering allele frequency for the variant.
+*while filtering allele frequency has been pre-computed for ExAC variants, the same framework can be readily applied using another reference sample.
 
@@
 Shown along with the frequency in cases is the estimated population allele frequency (calculated as: case frequency x disease prevalence x 1/2 x 1/variant penetrance) and the observed frequency in the ExAC dataset.
-^\*^As penetrance estimates for individual variants are not widely available, we have applied an estimate of 0.5 across all disorders (see **Supplementary Information**).
+^\*^As penetrance estimates for individual variants are not widely available, we have applied an estimate of 0.5 across these cardiac disorders (see **Supplementary Information**).
 HCM - hypertrophic cardiomyopathy; DCM - dilated cardiomyopathy; ARVC - arrhythmogenic right ventricular cardiomyopathy; LQTS - long QT syndrome.

### 2016-11-02 manuscript/SupplementaryMaterial.Rmd
Reducing word count

@@
 
+### Analysis tools
+
+R was used for all analyses, as well as to compile the figures, manuscript and supplementary material, making use of the following packages:
+knitr/rmarkdown (https://github.com/yihui/knitr)
+knitcitations (http://www.carlboettiger.info/2012/05/30/knitcitations.html)
+knitauthors (https://github.com/jamesware/knitauthors)
+dplyr (https://github.com/hadley/dplyr)
+ggplot2 (http://ggplot2.org/)
+packrat (https://rstudio.github.io/packrat/)
+
 ## Supplementary Table S1

### 2016-11-02 manuscript/manuscript.Rmd
Reducing word count

@@
 A variant's low frequency in, or absence from, reference databases is recognised as a necessary, but not sufficient, criterion for variant pathogenicity `r citep(c("10.1038/gim.2015.30","10.1038/nature13127"))`.
-The recent availability of very large reference databases, such as the Exome Aggregation Consortium (ExAC) `r citep("10.1038/nature19057")` dataset, which has characterised the population allele frequencies of 10 million genomic variants through the analysis of exome sequencing data from over 60,000 humans, provides an opportunity to obtain robust frequency estimates even for rare variants, improving the theoretical power for allele frequency (AF) filtering in Mendelian variant discovery efforts.
+The recent availability of very large reference databases, such as the Exome Aggregation Consortium (ExAC) `r citep("10.1038/nature19057")` dataset, which has characterised the population allele frequencies of 10 million genomic variants through analysis of exome sequencing data from over 60,000 humans, provides an opportunity to obtain robust frequency estimates even for rare variants, improving the theoretical power for allele frequency (AF) filtering in Mendelian variant discovery efforts.
 
@@
 In addition, for rare variants, estimation of true population AF is clouded by considerable sampling variance, even in the largest samples currently available.
-These limitations have encouraged the adoption of very lenient approaches when filtering variants by AF `r citep(c("10.1016/S0140-6736(14)61705-0","10.1038/ng.3304"))`, and recognition that more stringent approaches that account for disease-specific genetic architecture are urgently needed`r citep("10.1016/j.ajhg.2016.03.024")`.
+These limitations have encouraged the adoption of very lenient AF filtering approaches `r citep(c("10.1016/S0140-6736(14)61705-0","10.1038/ng.3304"))`, and recognition that more stringent approaches that account for disease-specific genetic architecture are urgently needed`r citep("10.1016/j.ajhg.2016.03.024")`.
 
-Here we present a statistical framework for assessing whether rare variants are sufficiently rare to cause penetrant Mendelian disease, while accounting for both architecture and sampling variance in observed allele counts.
-We demonstrate that AF cutoffs well below 0.1% are justified for a variety of human disease phenotypes and that such filters can remove an additional two-thirds of variants from consideration when compared to traditionally lenient frequency cutoffs, without discarding true pathogenic variants.
-We present pre-computed AF filtering values for all variants in the Exome Aggregation Consortium database, for comparison with user-defined disease-specific thresholds, which are now available through the ExAC data browser and for download, to assist others in applying this framework.
+Here we present a statistical framework for assessing whether variants are sufficiently rare to cause penetrant Mendelian disease, while accounting for both architecture and sampling variance in observed allele counts.
+We demonstrate that AF cutoffs well below 0.1% are justified for a variety of human disease phenotypes and that such filters can remove an additional two-thirds of variants from consideration compared to traditionally lenient frequency cutoffs, without discarding true pathogenic variants.
+We present pre-computed AF filtering values for all variants in the ExAC database, for comparison with user-defined disease-specific thresholds, which are available through the ExAC data browser and for download, to assist others in applying this framework.
 
@@
 We do not know the true population AF of any variant, having only an observed AF in a finite population sample.
-Moreover, confidence intervals around this observed frequency are problematic to estimate given our incomplete knowledge of the frequency spectrum of rare variants, which appears to be skewed towards very rare variants.
+Moreover, confidence intervals around this observed frequency are problematic to estimate given our incomplete knowledge of the frequency spectrum of rare variants, which is skewed towards very rare variants.
 For instance, a variant observed only once in a sample of 10,000 chromosomes is much more likely to have a frequency < 1:10,000 than a frequency >1:10,000.
@@
 
-To address this, we begin by specifying a maximum _true_ AF value we are willing to consider in the population (using the equation above), from which we can estimate the probability distribution for allele counts in a given sample size.
-This follows a binomial distribution, and can be satisfactorily approximated with a Poisson distribution (see **Supplementary Methods**).
+To address this, we begin by specifying a maximum _true_ AF value we are willing to consider in the population (using the equation above), from which we can estimate the probability distribution for allele counts in a given sample size (see **Supplementary Methods**).
 This allows us to set an upper limit on the number of alleles in a sample that is consistent with a given underlying population frequency.
@@
 
-To apply this threshold while remaining robust to chance variation in observed allele counts, we ask how many times a variant with population AF of 4.0x10^-5^ can be observed in a random population sample of a given size.
-For a 5% error rate we take the 95th percentile of a poisson distribution with λ = expected allele count, which is given by: sample size (chromosomes) &times; expected population AF (**Methods**).
-For HCM this gives us a maximum tolerated allele count of 9, assuming 50% penetrance (or 5 for fully penetrant alleles), for variants genotyped in the full ExAC cohort (sample size=121,412 chromosomes).
-The _MYBPC3_:c.1504C>T variant is observed 3 times in ExAC (freq=2.49x10^-5^; Table 1).
+To apply this threshold while remaining robust to chance variation in observed allele counts, we ask how many times a variant with population AF of 4.0x10^-5^ can be observed in a random population sample (**Methods**).
+At a 5% error rate, this yields a maximum tolerated allele count of 9, assuming 50% penetrance (5 for fully penetrant alleles) for variants genotyped in the full ExAC cohort (sample size=121,412 chromosomes).
+The _MYBPC3_:c.1504C>T variant is observed 3 times in ExAC (freq=2.49x10^-5^;Table 1).
 
@@
 
-All data required to reproduce these analyses is available at <https://github.com/ImperialCardioGenetics/frequencyFilter>.
-The manuscript was compiled in R, and source code for the analysis, figures and manuscript, are available at the same location.
-Curated variant interpretations are deposited in ClinVar *Accession & DOI to be added*.
+All data required to reproduce the analysis, figures and manuscript (compiled in R) are available at <https://github.com/ImperialCardioGenetics/frequencyFilter>.
+Curated variant interpretations are deposited in ClinVar *Details to be added*.
 ExAC annotations are available at <https://exac.broadinstitute.org>.
@@
 
-```{r, echo=FALSE, message=FALSE}
-write.bibtex(file="references.bib")
-```
-
 ##FIGURE AND TABLE LEGENDS
@@
 ### Figure 1
-> Plot of ExAC allele count (all populations) against case allele count for variants classified as VUS, Likely Pathogenic or Pathogenic in 6179 HCM cases.
+Plot of ExAC allele count (all populations) against case allele count for variants classified as VUS, Likely Pathogenic or Pathogenic in 6179 HCM cases.
 The dotted lines represent the maximum tolerated ExAC allele counts in HCM for 50% (dark blue) and 100% penetrance (light blue).
@@
 
-> The clinical utility of stringent allele frequency thresholds.
+The clinical utility of stringent allele frequency thresholds.
 (a) The number of predicted protein-altering variants (definition in Methods) per exome as a function of the allele frequency filter applied.
@@
 
-> A flow diagram of our approach, applied to a dominant condition, and using ExAC as our reference sample.
+A flow diagram of our approach, applied to a dominant condition, and using ExAC as our reference sample.
 First, a disease-level maximum credible population allele frequency is calculated, based on disease prevalence, heterogeneity and penetrance.
@@
 ### Table 1  
-> Details of the most prevalent pathogenic variants in case cohorts for five cardiac conditions.
+Details of the most prevalent pathogenic variants in case cohorts for five cardiac conditions.
 Shown along with the frequency in cases is the estimated population allele frequency (calculated as: case frequency x disease prevalence x 1/2 x 1/variant penetrance) and the observed frequency in the ExAC dataset.
@@
 ### Table 2
-> Maximum credible population frequencies and maximum tolerated ExAC allele counts for variants causative of exemplar inherited cardiac conditions, assuming a penetrance of 0.5 throughout.
+Maximum credible population frequencies and maximum tolerated ExAC allele counts for variants causative of exemplar inherited cardiac conditions, assuming a penetrance of 0.5 throughout.
 CPVT - catecholaminergic polymorphic ventricular tachycardia; FH - familial hypercholesterolaemia.
 Prevalence estimates (taken as the highest value reported) were obtained from: Marfan`r citep("10.1016/S0140-6736(05)67789-6")`, Noonan`r citep("10.1016/S0140-6736(12)61023-X")`, CPVT`r citep("10.1038/ejhg.2013.55")` and classical Ehlers-Danlos`r citep("10.1097/GIM.0b013e3181eed412")`.
+
+```{r, echo=FALSE, message=FALSE}
+write.bibtex(file="references.bib")
+```
 

### 2016-11-02 R/Figures.R
Changing figure numbers

@@ -2,7 +2,7 @@
 #source("../R/setup.R")
 my.CI<-0.95
 
-### Figure 2
+### Figure 1
 
 variants<-read.table("../data-raw/LMM_OXF_path_likelyPath_VUS_variants_allExACpopulations_withLowestClass.txt", header=TRUE, sep="\t")
 maxHCMPenTerm<-9
@@ -39,7 +39,7 @@ zoomplot <- ggplot(subset(variants, (Disease=="HCM" & !(Gene %in% c('GLA','DMD',
         legend.position="none",  axis.line.x = element_line(color="grey"), axis.line.y = element_line(color="grey"), panel.background = element_blank(), 
         axis.ticks = element_line(color="grey"))
 
-png("../figures/Figure2.png",width=1200,height=1000,res=150)
+png("../figures/Figure1.png",width=1200,height=1000,res=150)
 subvp<-viewport(width=.5,height=.5,x=.745,y=.745)
 
 zoomplot
@@ -47,7 +47,7 @@ print(zoomplot)
 print(fullplot,vp=subvp)
 dev.off()
 
-### Figure 3
+### Figure 2
 
 # Details for Exome plot (figure 3a)
 stats = read.table('../data/filtering_stats.tsv',sep='\t',quote='',header=TRUE)
@@ -121,7 +121,7 @@ genes = c(names(gene_names)[c(1,2,3)])
 ef_data$acol = gene_colors[ef_data$Gene]
 
 # Plot multipanel plot
-png('../figures/Figure3.png',width=3800,height=1500,res=150)
+png('../figures/Figure2.png',width=3800,height=1500,res=150)
 m<-rbind(c(1,2))
 layout(m)
 

### 2016-12-16 R/Figures.R
Altering odds ratio figure in response to reviewers comments to take internal HVOL data

@@ -39,7 +39,7 @@ zoomplot <- ggplot(subset(variants, (Disease=="HCM" & !(Gene %in% c('GLA','DMD',
         legend.position="none",  axis.line.x = element_line(color="grey"), axis.line.y = element_line(color="grey"), panel.background = element_blank(), 
         axis.ticks = element_line(color="grey"))
 
-png("../figures/Figure1.png",width=1200,height=1000,res=150)
+png("../figures/Figure1.png",width=2400,height=2000,res=300)
 subvp<-viewport(width=.5,height=.5,x=.745,y=.745)
 
 zoomplot
@@ -71,40 +71,40 @@ stats$acol = pop_colors[stats$ancestry]
 # Compute Odds Ratios and Confidence Intervals from ExAC and Internal HCM data
 #install.packages("tidyr")
 options(stringsAsFactors = FALSE)
-exac_data<-read.table("../data-raw/ExACVariants_0.001_sarcomeric.txt", header=TRUE, sep="\t") %>% 
-  tbl_df %>%
-  transmute(Gene,CodingVar,AC=ExACAlleleCount,exacAF=ExACFreq,group="control")
+#exac_data<-read.table("../data-raw/ExACVariants_0.001_sarcomeric.txt", header=TRUE, sep="\t") %>% 
+#  tbl_df %>%
+#  transmute(Gene,CodingVar,AC=ExACAlleleCount,exacAF=ExACFreq,group="control")
 hcm_data<-read.table("../data-raw/HCM_BRU_Variants_and_freqs.txt", header=TRUE, sep="\t") %>% 
   tbl_df %>%
   transmute(Gene,CodingVar,AC=CaseCount,exacAF=ExACFreq,group="case")
+hvol_data<-read.table("../data-raw/HVOL_BRU_Variants_and_freqs.txt", header=TRUE, sep="\t") %>% 
+  tbl_df %>%
+  transmute(Gene,CodingVar,AC=CaseCount,exacAF=ExACFreq,group="control")
 
-data <- rbind(exac_data,hcm_data)
-data$bin <- cut(data$exacAF,breaks=c(1,0.001,0.0005,0.0001,0.00005,0.00001,-1),labels=c("0.00001","0.00005","0.0001","0.0005","0.001","1"))
[…]

### 2016-12-16 manuscript/SupplementaryMaterial.Rmd
Responding to reviewers comments

--- 2016-11-01 manuscript/SupplementaryMaterial.Rmd
@@
 
+For dominant disorders we define the maximum credible population AF (for a pathogenic allele) as:
+
+*maximum credible population AF = prevalence* $\times$ *maximum allelic contribution* $\times$ *1/penetrance*
+
+where _maximum allelic contribution_ is the maximum proportion of cases potentially attributable to a single allele, a measure of heterogeneity.
+It is important to note that although we estimate the _maximum allelic contribution_ as a single value, this can be broken down into seperate factors corresponding to the maximum contribution of any one gene (_maximum genetic contribution_) and the _maximum allelic contribution_ within that gene.
+These two seperate factors are used in the derivation of the recessive model described below.
+
 The maximum tolerated allele count (AC) was computed as the AC occurring at the upper bound of the one-tailed 95% confidence interval (95%CI AC) for the established maximum credible allele frequency, given the observed allele number (AN).
@@
 
-The prevalence of a recessive condition can be related to the AF of causative variants by:  
-*Prevalence =* $\sum$ *(AF of causative alleles in each contributing gene)^2^ x penetrance*  
-approximating to:   
-*Prevalence = (combined frequency of causative alleles in gene)^2^ x (number of similar genes) x penetrance*  
-and expanding to:  
-*Prevalence = (max individual AF x 1/maximum allelic contribution)^2^ x 1/maximum genetic contribution x penetrance*
+For a biallelic condition caused by only a single gene, the prevalence can be determined from the combined frequency of all possible pairs of causative alleles in that gene:
 
-where *maximum genetic contribution* represents the proportion of all cases that are attributable to the gene under evaluation, and *maximum allelic contribution* represents the proportion of cases attributable to that gene that are attributable to an individual variant.
-This can be rearranged to give the maximum frequency of a recessive disease causing variant in the population as:
-*max credible AF* = $\sqrt($*prevalence*$) \times$ *maximum allelic contribution* $\times \sqrt($*maximum genetic contribution*$) \times 1/\sqrt($*penetrance*$)$
+$prevalence = \sum_{i}^{j} (AF_{i} \times AF_{j} \times penetrance_{i,j})$
+
+where $AF_{i}$ and $AF_{j}$ represent the frequencies of the two alleles, and $i$ and $j$ may be the same allele.
+
+For a condition caused by multiple genes, the above equation must then be summed across $n$ contributing genes:
+
+$prevalence = \sum_{1}^{n} \sum_{i}^{j} (AF_{i} \times AF_{j} \times penetrance_{i,j})$
+
+Since the penetrance can not be separately estimated for each combination of alleles (though is typically high for recessive conditions), we simplify by estimating penetrance as constant for the condition, reducing the equation to:
+
+$prevalence = penetrance \times \sum_{1}^{n}\sum_{i}^{j} (AF_{i} \times AF_{j})$
+
+which is equivalent to:
+
+$prevalence = penetrance \times \sum_{1}^{n} \sum_{i}^{} (AF_{i}^2)$
+
+we can rearrange this to give an upper bound for the maximum credible AF of an individual causative variant:
+
+$max$ $credible$ $AF$ = $\sqrt(prevalence) \times$ $maximum$ $allelic$ $contribution$ $\times \sqrt(maximum$ $genetic$ $contribution$) $\times 1/\sqrt(penetrance)$
+
+where rather than simply dividing by n genes to get an average, we take the upper bound of the proportion attributable to any single gene (the *maximum genetic contribution*).
+Similarly we take the maximum proportion of cases attributable to that gene that are attributable to any individual variant (the *maximum allelic contribution*).
 
@@
 We believe that the ease of re-calculating our "maximum credible population allele frequency" lends itself to any of these approaches.
-We provide an online calculator to facilitate the exploration of these parameters (<https://jamesware.shinyapps.io/alleleFrequencyApp/>).
+We provide an online calculator to facilitate the exploration of these parameters (<http://cardiodb.org/alleleFrequencyApp>).
 If there are large case and control populations for a disease and the diease prevalence is known, we can use these to estimate penetrance `r citep("10.1126/scitranslmed.aad5169")`.
@@
 Similarly, because we have not assigned ancestry for the "Other" individuals, it is difficult to assess the population frequency of variants seen only in this set of individuals.
-Users are left to judge whether variants that  would not be filtered on the basis of frequency in the five continental populations, but that are recurrent in Finnish or "Other" populations, should be removed from consideration according to the specific circumstances.
+Users are left to judge whether variants that  would not be filtered on the basis of frequency in the five continental populations, but that are sufficiently frequent in Finnish or "Other" populations, should be removed from consideration according to the specific circumstances.
 
@@
 
+### Analysis tools
+
+R was used for all analyses, as well as to compile the figures, manuscript and supplementary material, making use of the following packages:
+knitr/rmarkdown (https://github.com/yihui/knitr), knitcitations (http://www.carlboettiger.info/2012/05/30/knitcitations.html), knitauthors (https://github.com/jamesware/knitauthors), dplyr (https://github.com/hadley/dplyr), ggplot2 (http://ggplot2.org/), packrat (https://rstudio.github.io/packrat/).
+
 ## Supplementary Table S1
@@
 Variants previously reported as causative of HCM either in ClinVar, or in a clinical series of 6179 HCM cases, but that were observed in ExAC above the maximum tolerated allele count for HCM (AC>9 globally) were manually curated according to the ACMG guidelines for interpretation and classification on sequence variants.
+For 5/43 variants, ClinVar contained the only identifiable disease-associated variant record, while 38/43 had additional information in the literature.
+Curated variant interpretations are deposited in ClinVar under the submission name "HCM\_ExAC\_frequency\_review\_2016".
 *These variants were previously categorised as Likely Pathogenic in the LMM and Oxford HCM cohorts.

### 2016-12-16 manuscript/manuscript.Rmd
Responding to reviewers comments

@@
 read.delim("../data-raw/authors.txt") %>%
-  formatAuthors
+  formatAuthors(infoText2="These authors jointly supervised this work")
 ```
@@
 Disease prevalence estimates were obtained from the literature and taken as the highest value reported.
-Cardiovascular diseases variants were modelled with a pentrance of 0.5, corresponding to the reported penetrance of the HCM variant used to illustrate our approach `r citep("10.1161/CIRCRESAHA.109.216291")` and the minimum found accross a range of variants/disorders.
+Cardiovascular disease variants were modeled with a penetrance of 0.5, corresponding to the reported penetrance of the HCM variant used to illustrate our approach `r citep("10.1161/CIRCRESAHA.109.216291")` and the minimum found across a range of variants/disorders.
 
@@
 ExAC counts were determined by matching on chromosome, position, reference, and alternate alleles.
-For all variants above the proposed maximum tolerated allele count for HCM, HGMD literature was reviewed and the level of evidence supporting pathogenicity was curated according to ACMG criteria `r citep("10.1038/gim.2015.30")`.
+For all variants above the proposed maximum tolerated allele count for HCM, literature from both HGMD and PubMed was reviewed and the level of evidence supporting pathogenicity was curated according to ACMG criteria `r citep("10.1038/gim.2015.30")`.
 
@@
 
-322 HCM patients (confirmed by cardiac MRI) recruited to the NIHR Royal Brompton cardiovascular BRU were sequenced using the IlluminaTruSight Cardio Sequencing Kit `r citep("10.1007/s12265-016-9673-5")` on Illumina MiSeq and NextSeq platforms.
+322 HCM patients and 852 healthy volunteers (both confirmed by cardiac MRI) recruited to the NIHR Royal Brompton cardiovascular BRU were sequenced using the IlluminaTruSight Cardio Sequencing Kit `r citep("10.1007/s12265-016-9673-5")` on Illumina MiSeq and NextSeq platforms.
 This study had ethical approval (REC: 09/H0504/104+5) and informed consent was obtained for all subjects.
-The number of rare variants in _MYBPC3_, _MYH7_ and the six other sarcomeric genes associated with HCM (_TNNT2_, _TNNI3_, _MYL2_, _MYL3_, _TPM1_ and _ACTC1_) were calculated for this HCM cohort, and for reference population samples from ExAC.
-Variant frequencies were calculated for all protein altering variants (frameshift, nonsense, splice donor/acceptor, missense and in-frame insertions/deletions), with case/control odds ratios calculated separately for non-overlapping ExAC AF bins with the following breakpoints: 1x10^-5^, 5x10^-5^, 1x10^-4^, 5x10^-4^ and 1x10^-3^.
-Odds Ratios were calculated as OR = (cases with variant/cases without variant) / (ExAC samples with variant/ExAC samples without variant).
-In the absence of sample-level genotype data for ExAC, the number of samples with a variant was approximated by the total number of variant alleles (assuming each rare variant was from a distinct sample).
+The number of rare variants in the eight sarcomeric genes associated with HCM (_MYBPC3_, _MYH7_, _TNNT2_, _TNNI3_, _MYL2_, _MYL3_, _TPM1_ and _ACTC1_) were calculated for all protein altering variants (frameshift, nonsense, splice donor/acceptor, missense and in-frame insertions/deletions), with case/control odds ratios calculated separately for non-overlapping ExAC AF bins with the following breakpoints: 4x10^-5^, 1x10^-4^, 5x10^-4^ and 1x10^-3^.
+Odds Ratios were calculated as OR=(cases with variant/cases without variant)/(controls with variant/controls without variant).
 
@@
 
-To facilitate these calculations, we have produced an online calculator (<https://jamesware.shinyapps.io/alleleFrequencyApp/>) that will compute maximum credible population AF and maximum sample allele count for a user-specified genetic architecture, and conversely allow users to dynamically explore what genetic architecture(s) might be most compatible with an observed variant having a causal role in disease.
+To facilitate these calculations, we have produced an online calculator (<http://cardiodb.org/alleleFrequencyApp>) that will compute maximum credible population AF and maximum sample allele count for a user-specified genetic architecture, and conversely allow users to dynamically explore what genetic architecture(s) might be most compatible with an observed variant having a causal role in disease.
 
-#### *Assessing the acuracy of our approach*
+#### *Assessing the accuracy of our approach*
 
-For all diseases with cases series that permitted us to define the genetic arcitecture, the commonest variant in the cases series was well within the calculated maximum alelle count in ExAC (Table 1).
+For all diseases with cases series that permitted us to define the genetic architecture, the commonest variant in the cases series was well within the calculated maximum allele count in ExAC (Table 1).
 
@@
 For example, a variant relatively common in any one population is unlikely pathogenic, even if rare in other populations, provided the disease prevalence and architecture is consistent across populations.
-We therefore compute a maximum tolerated AC for each distinct sub-population of our reference sample, and filter based on the highest AF observed in any major continental population (see **Methods** and **Suplementary Note 3**).
+We therefore compute a maximum tolerated AC for each distinct sub-population of our reference sample, and filter based on the highest AF observed in any major continental population (see **Methods**).The tightness of the Poisson distribution used to compute maximum tolerated allele count is a function of sample size, and thus our approach is more conservative when allele number is lower, thus avoiding inappropriately filtering variants due to chance observation of a few alleles in a smaller sub-population or at a poorly genotyped site (see **Supplementary Note 3**).
 
@@
 
-Finally, if the allelic heterogeneity of a disorder is not well characterised, it is conservative to assume minimal heterogeneity, so that the contribution of each gene is modelled as attributable to one allele, and the maximum allelic contribution is substituted by the maximum genetic contribution (i.e the maximum proportion of the disease attributable to single gene).
+Finally, if the allelic heterogeneity of a disorder is not well characterised, it is conservative to assume minimal heterogeneity, so that the contribution of each gene is modeled as attributable to one allele, and the maximum allelic contribution is substituted by the maximum genetic contribution (i.e the maximum proportion of the disease attributable to single gene).
 For classic Ehlers-Danlos syndrome, up to 40% of the disease is caused by variation in the _COL5A1_ gene `r citep("10.1097/GIM.0b013e3181eed412")`.
@@
 
-Our method is designed to be complemenary to and used along side other gene and variant level methods to filter and prioritise candiate variants (e.g.
-gene level constraint`r citep("10.1038/ng.3050")`, amino acid conservation and missense prediction algorithms) along with segregation and functional data.
+Our method is designed to be complementary to and used along side other gene and variant level methods to filter and prioritise candiate variants (e.g.
+gene level constraint`r citep("10.1038/ng.3050")`, amino acid conservation`r citep(c("10.1101/gr.3715005","10.1101/gr.3577405"))` and missense prediction algorithms`r citep(c("10.1038/nmeth0410-248","10.1093/nar/gks539"))`) along with segregation and functional data.
 
@@
 While incomplete knowledge of the genetic architecture of a disease of interest will limit this or any approach to evaluate a specific variant that has been observed at low frequency in a reference population, our framework and accompanying web tool do at least transparently define the range of disease architectures that are compatible with the observed data.
-For example, we may not be able to say definitively that a variant is benign, but we might be confident that it could only be causal if the prevalence of the disease is higher than published estimates, or its penetrance is below a specified value.
+For example, many neurological disorders have Mendelian forms as well as idiopathic forms with genetic risk factors of modest effect sizes, high allelic and genetic heterogeneity, and/or dramatic variability in the penetrance of different variants `r citep(c("10.1016/j.cell.2013.10.037","10.1016/j.cell.2015.01.019","10.1016/j.cell.2015.01.052","10.1371/journal.pgen.1002548","10.1126/scitranslmed.aad5169"))`.
+Reference population allele frequency information alone can never definitively show that a variant possesses no association to disease, but it can still provide sensible constraints.
+The calculations described here can be used to show that a variant could could only be causal if the prevalence of the disease is higher than published estimates, or its penetrance is below a specified value `r citep("10.1016/j.neuron.2016.09.054")`.
 
@@
 All data required to reproduce the analysis, figures and manuscript (compiled in R) are available at <https://github.com/ImperialCardioGenetics/frequencyFilter>.
-Curated variant interpretations are deposited in ClinVar *Details to be added*.
+Curated variant interpretations are deposited in ClinVar under the submission name "HCM\_ExAC\_frequency\_review\_2016".
 ExAC annotations are available at <https://exac.broadinstitute.org>.
-Our AF calculator app is located at <https://jamesware.shinyapps.io/alleleFrequencyApp/>, with source code available at <http://github.com/jamesware/alleleFrequencyApp>.
+Our AF calculator app is located at <http://cardiodb.org/alleleFrequencyApp>, with source code available at <http://github.com/jamesware/alleleFrequencyApp>.
 
@@
 (b) The odds ratio for HCM disease-association against allele frequency.
-The prevalence of variants in HCM-associated genes (_MYH7_, _MYBPC3_ and other sarcomeric  (_TNNT2_, _TNNI3_, _MYL2_, _MYL3_, _TPM1_ and _ACTC1_, analysed collectively)) in 322 HCM cases and 60,706 ExAC controls were compared for a range of allele frequency bins, and an odds ratio computed (see **Methods**).
+The prevalence of variants in sarcomeric HCM-associated genes (_MYH7_, _MYBPC3_, _TNNT2_, _TNNI3_, _MYL2_, _MYL3_, _TPM1_ and _ACTC1_, analysed collectively) in 322 HCM cases and 852 healthy controls were compared for a range of allele frequency bins, and an odds ratio computed (see **Methods**).
 Data for each bin is plotted at the upper allele frequency cutoff.

### 2017-01-19 manuscript/SupplementaryMaterial.Rmd
Changes to recessive derivation

@@
 
-For a biallelic condition caused by only a single gene, the prevalence can be determined from the combined frequency of all possible pairs of causative alleles in that gene:
+For a biallelic condition caused by only a single gene, the prevalence can be determined from the combined frequency of all possible pairs of alleles in that gene:
 
-$prevalence = \sum_{i}^{j} (AF_{i} \times AF_{j} \times penetrance_{i,j})$
+$prevalence = \sum_{i,j} (AF_i \times AF_j \times penetrance_{i,j})$
 
-where $AF_{i}$ and $AF_{j}$ represent the frequencies of the two alleles, and $i$ and $j$ may be the same allele.
+where $penetrance_{i,j}$ is the penetrance of two alleles $i$ and $j$ with frequencies $AF_{i}$ and $AF_{j}$ and $i$ and $j$ may be the same allele.
+This is generalisable in that all variants may be included in the summation, as the $penetrance_{i,j}$ will be 0 for benign variants, and those variants will not contribute to disease.
 
@@
 
-$prevalence = \sum_{1}^{n} \sum_{i}^{j} (AF_{i} \times AF_{j} \times penetrance_{i,j})$
+$prevalence = \sum_{1}^{n} \sum_{i,j} (AF_{i} \times AF_{j} \times penetrance_{i,j})$
 
-Since the penetrance can not be separately estimated for each combination of alleles (though is typically high for recessive conditions), we simplify by estimating penetrance as constant for the condition, reducing the equation to:
+where $penetrance_{i,j}$ is the penetrance of $i$ and $j$.
 
-$prevalence = penetrance \times \sum_{1}^{n}\sum_{i}^{j} (AF_{i} \times AF_{j})$
+Since the penetrance cannot be separately estimated for each combination of alleles (though is typically high for recessive conditions), we simplify by estimating penetrance as constant for the condition, and calculating only for pathogenic alleles, reducing the equation to:
 
-which is equivalent to:
+$prevalence = penetrance \times \sum_{1}^{n}\sum_{i,j} (AF_{i} \times AF_{j})$
 
-$prevalence = penetrance \times \sum_{1}^{n} \sum_{i}^{} (AF_{i}^2)$
+where $penetrance$ is the penetrance for any pair of pathogenic alleles and $i$ and $j$ are both pathogenic alleles.
 
-we can rearrange this to give an upper bound for the maximum credible AF of an individual causative variant:
+Since we are now treating penetrance as constant for each gene this simplifies further: for a given gene containing $m$ pathogenic alleles, the frequency of individuals who are homozygous or compound heterozygous is given by the square of the combined allele frequencies of all contributing alleles.
+
+$\sum_{i,j} (AF_{i} \times AF_{j}) = (\sum_{i=1}^{m}AF_{i})^2$
+
+If we know the AF of a single variant $i$, and the contribution that variant makes to disease, then the combined allele frequencies of all contributing alleles ($\sum_{i=1}^{m}AF_{i}$) could in turn be represented as $\frac {AF_{i}}{allelicContribution_{i}}$ where $allelicContribution_{i}$ is the proportion of disease that is attributable to variant $i$.
+
+Substituting this into our equation for prevalence yields:
+
+$prevalence = penetrance \times \sum_{1}^{n}(\frac {AF_{i}}{allelicContribution_{i}})^2$
+
+For a specific variant in a specific gene, with known allelic and genetic contribution (i.e.
+the proportion of cases attributable to the gene that contains variant $i$), prevalence can be expressed as:
+
+$prevalence = penetrance \times (\frac {1}{geneticContribution_{i}}) \times (\frac {AF_{i}}{allelicContribution_{i}})^2$
+
+Finally, we can rearrange this to give an upper bound for the maximum credible AF of an individual causative variant:
 
@@
 
-where rather than simply dividing by n genes to get an average, we take the upper bound of the proportion attributable to any single gene (the *maximum genetic contribution*).
-Similarly we take the maximum proportion of cases attributable to that gene that are attributable to any individual variant (the *maximum allelic contribution*).
+where $maximum$ $genetic$ $contribution$ is the maximum proportion of cases attributatble to any one gene and $maximum$ $allelic$ $contribution$ is the maximum proportion of cases attributable to any one variant within that gene.
 
